Nutritional state-dependent modulation of Insulin-Producing Cells in Drosophila

Version 1

Abstract

Insulin plays a key role in regulating metabolic homeostasis across vertebrate and invertebrate species. Drosophila Insulin-Producing Cells (IPCs) are functional analogues to mammalian pancreatic beta cells and release insulin directly into circulation. IPC activity is modulated by nutrient availability, circadian time, and the behavioral state of animals. To investigate the in vivo dynamics of IPC activity in the context of metabolic homeostasis, we quantified effects of nutritional and internal state changes on IPCs using electrophysiological recordings. We found that the nutritional state strongly modulates IPC activity. IPCs were less active in starved flies than in fed flies. Refeeding starved flies with glucose significantly increased IPC activity, suggesting that IPCs are regulated by hemolymph sugar levels. In contrast to glucose feeding, glucose perfusion had no effect on IPC activity. This was reminiscent of the mammalian incretin effect, in which ingestion of glucose drives higher insulin release than intravenous glucose application. Contrary to IPCs, Diuretic hormone 44-expressing neurons in the pars intercerebralis (DH44PINs), which are anatomically similar to IPCs, significantly increased their activity during glucose perfusion. Functional connectivity experiments based on optogenetic activation demonstrated that glucose-sensing DH44PINs do not affect IPC activity, while other DH44Ns inhibit IPCs. This suggests that populations of autonomously and systemically glucose-sensing neurons are working in parallel to maintain metabolic homeostasis. Ultimately, metabolic state changes affect animal behavior. For example, hungry flies increase their locomotor activity in search of food to maintain metabolic homeostasis. In support of this idea, activating IPCs had a small, satiety-like effect in starved flies, resulting in reduced walking activity, whereas activating DH44Ns strongly increased walking activity. Taken together, we show that IPCs and DH44Ns are an integral part of a sophisticated modulatory network that orchestrates glucose homeostasis and adaptive behavior in response to shifts in the metabolic state.

Introduction

Insulin is a key neuropeptide that controls metabolic homeostasis across the animal kingdom1–4. Beyond metabolic homeostasis, insulin signaling is implicated in physiological processes underlying reproduction, aging, and stress resistance5–7. When nutrients are abundant, insulin promotes the uptake and storage of energy. Conversely, during periods of starvation, insulin release is downregulated8,9, which promotes the conservation of energy resources and enables energy mobilization via counter-regulatory pathways10–12. Maintaining this energy balance is vital for survival. Accordingly, dysregulation of this finely tuned system is a hallmark of metabolic disorders such as type 2 diabetes13–15.

Drosophila melanogaster has emerged as a model system to study insulin signaling because of the availability of a powerful genetic toolkit16–21. In Drosophila, Insulin-Producing Cells (IPCs), which are analogous to human pancreatic beta cells, reside in the pars intercerebralis (PI) of the brain22. These IPCs secrete Drosophila insulin-like peptides (DILPs) and are sensitive to changes in the metabolic state23–26, the behavioral state27,28, the circadian time25,29, aminergic and peptidergic inputs30, and other peripheral signals31. Consequently, they play a key role in orchestrating the metabolic demands of behaving animals. For instance, insulin acts as a global cue for modulating odor-driven food-search in accordance with a fly’s nutritional state32. Unlike pancreatic beta cells, IPCs are localized in the brain and engage in direct synaptic signalling33 in addition to endocrine signalling. Thus, IPCs are part of a complex system that integrates both neuronal and peripheral signals. As a key hub integrating metabolic state information, IPCs are thought to be part of the neuronal circuitry that regulates food intake32,34–39, in part by regulating foraging behavior. Previous studies found that octopamine, the insect analogue to norepinephrine, mediates starvation-induced hyperactivity in flies38. A subset of octopaminergic neurons expressing Drosophila insulin receptor (dInR) are necessary to regulate starvation-induced hyperactivity38,40 implicating that IPCs play a role in modulating foraging behavior.

While much is known about carbohydrate metabolism and insulin signaling in Drosophila5,28,41,42, the in vivo dynamics of IPCs remain largely elusive. Ex vivo studies indicate that IPCs themselves are sensitive to glucose, as evidenced by increased firing rates upon glucose application23,43. However, the influence of feeding and nutritional states on IPC activity in vivo, as well as the nutritional state-dependent response of IPCs to glucose, remain unclear, underscoring the importance of in vivo studies where all nutrient-sensing and sensorimotor circuits are intact. For instance, how does IPC activity differ when glucose is ingested as part of a meal, compared to when glucose is administered directly to the brain or into circulation? In mammalian systems, insulin release is significantly amplified upon glucose ingestion vs. intravenous administration44,45, a phenomenon known as the ‘incretin effect’ 46,47. This underlines the complexities of insulin regulation, emphasizing that insulin secretion is not solely dictated by sugar levels but influenced by multiple factors. Here, we investigated the nutritional state-dependent modulation of IPCs in Drosophila using an in vivo electrophysiological approach.

We recently demonstrated that IPC activity is strongly modulated by locomotion on fast timescales27. Whether the converse is true, and IPC activity can directly affect behavior on short timescales, is not clear. To address this, we optogenetically manipulated IPCs and other modulatory neurons while recording the locomotor activity of freely walking flies.

Overall, we demonstrate that IPC activity is not just nutritional state-dependent, but that IPCs also exhibit an incretin-like effect. We further establish that the nutritional state strongly modulates locomotor activity, and that IPCs play a small but significant role in this. In addition, we confirm that another set of modulatory neurons adjacent to the IPCs, DH44PINs, are sensitive to glucose perfusion. While DH44PINs do not affect IPC activity on short-timescales, DH44 neurons outside the PI (DH44Ns) form strong inhibitory connections to IPCs. These DH44Ns also have strong effects on locomotor activity, which are antagonistic to those of IPCs. Our findings imply distinct mechanisms by which IPCs and DH44PINs sense changes in nutritional state, underlining the significance of multiple pathways in orchestrating Drosophila’s response to nutritional state changes and ensuring metabolic homeostasis.

Results

IPC activity is modulated by the nutritional state

To characterize the electrophysiological activity of IPCs in different nutritional states, we performed in vivo patch-clamp recordings from IPCs in fed and starved Drosophila (Figure 1A-C). First, flies were fed ad libitum, and the IPC baseline activity was determined by recording their spontaneous activity in glucose-free extracellular saline over 10 minutes. The baseline firing rate varied between individual IPCs and ranged from 0 to 1.4 Hz (Figure 1D). To assess the effect of starvation on IPC activity, we wet-starved flies for 24 hours. IPCs in starved flies were significantly less active, and basically remained quiescent with 14 out of 16 IPCs recorded (88%) exhibiting a firing rate of 0 Hz (Figure 1D). We also analyzed the resting membrane potential (Vm) of IPCs in fed and starved flies. On average, the Vm of IPCs was 9 mV more depolarized (median: -50 mV) in fed flies compared to starved flies (median: -59 mV, Figure 1E). This indicates a higher excitability of IPCs after feeding, which likely explains their higher firing rates. These results demonstrate that IPC activity and hence insulin release are strongly modulated by the nutritional state.

IPC activity depends on the nutritional state and increases after glucose ingestion.

A) Schematic of the setup for in vivo IPC whole cell patch-clamp recordings. B) IPCs in the Drosophila brain. UAS-myr-GFP was expressed under a Dilp2-GAL4 driver to label IPCs. The GFP signal was enhanced with anti-GFP (green), brain neuropils were stained with anti-nc82 (magenta). C) Representative examples of the membrane potential of two IPCs recorded in fed (magenta) and starved (cyan) flies. D) Average baseline spike rate and E) membrane potential of IPCs in fed (magenta) and starved (cyan) flies. Each dot represents an individual IPC, error bars indicate median (circle) and inter-quartile range (IQR, bars). P-values from Wilcoxon rank-sum test. F) Schematic of the experimental starvation and refeeding protocol. HG: High glucose, HP: High protein, SD: Standard diet. G) Comparison of IPC spike rates in fed flies (magenta), 24 h starved flies (cyan), and flies refed on HG for different durations (green). H) Comparison of IPC spike rate in 24 h starved flies (cyan), HP (yellow) and SD (grey). Each circle represents an individual IPC, N = number of IPCs (at least 7 flies were used for each condition).

Since IPCs are central to metabolic regulation and development, factors other than the nutritional state may impact their activity. To address two of the key factors, we analyzed whether IPC activity was affected by the mating state or the age of the fly. The IPC activity in virgin flies was slightly lower than that of mated females, but this difference was not significant (Supplementary Figure S1A). Although IPC activity was highest in males, there were no significant differences between the three groups (Supplementary Figure S1A). Hence, IPC activity is not significantly affected by sex or female mating state. However, when recording the IPC activity in females of different age groups, we observed that IPCs remained almost quiescent in fully fed females older than ten days. Here, baseline spike rates ranged from 0 – 0.2 Hz and were significantly lower than in younger females (Supplementary Figure S1B). This indicates that IPC activity is suppressed in older flies. Therefore, aging significantly reduces IPC activity.

IPCs are sensitive to glucose ingestion

To investigate whether and how IPC activity is modulated by nutrient uptake in vivo, we starved flies for 24 h and then refed them with a high glucose (HG) diet (400 mM glucose in 2% agar) for durations between 3 to 24 h (Figure 1F). IPC activity in flies that were refed for 3-5 h resembled that of starved flies and remained around 0 Hz (Figure 1G). Six hours after the start of refeeding, IPC activity started to increase, and was indistinguishable from that of fed flies (Figure 1G, p = 0.68), indicating that this duration was sufficient to allow IPC activity to fully recover from starvation. Interestingly, IPC activity kept increasing for longer glucose feeding durations, and was significantly increased in flies that were refed with glucose for 18-24 h compared to flies kept on a full diet and fed ad libitum (Figure 1G, p = 0.02). This indicates that the release of insulin from IPCs increased when flies were exclusively feeding on glucose. We conclude that IPC activity was increased in flies feeding on a pure sugar diet, and that the temporal dynamics of this effect were relatively slow.

Next, we asked whether the increased IPC activity after refeeding was glucose-specific or a general response to food intake. To test this, we kept 24 h starved flies on either a standard (SD) or high protein diet (HP) and compared the IPC activity after refeeding for 18-24 h. In contrast to refeeding flies with HG diets, refeeding them with either HP or SD diets caused no significant increase in the spike rate of IPCs compared to starved flies (p = 0.43 and 0.77, respectively, Figure 1H). This was intriguing, because both protein-rich diets and full diets have been shown to induce insulin secretion under fed conditions26,48,49. To ensure that these results were not due to a lasting interference with the fly’s metabolic system caused by keeping the flies on an extreme diet, we conducted a survival assay. Analogous to our electrophysiology experiments, we starved the flies for 24 h and then refed them on HG, SP, or SD. Subsequently, we monitored the number of survivors every day while flipping flies onto fresh food every second day. As expected, flies fed with SD survived the longest (100 % survival until day 43, Supplementary Figure S1C). 100 % survival was observed for flies fed with HG until Day 12 and at least half of the flies survived until day 16 (Supplementary Figure S1C). In comparison, flies survived poorly on HP diet: 100 % survival was observed only up to day 1 and more than half of the flies were dead by day 12 (42.5 % survival, Supplementary Figure S1C). Since our IPC recordings were acquired in flies refed on HP or HG for a maximum of 24 hours, the effects we observed on IPC activity were not due to lethality in flies but rather due to differences in nutrient availability. Hence, we conclude that the increase in IPC activity after refeeding was indeed glucose specific.

The nutritional state affects locomotor activity

Starvation increases baseline locomotor activity in flies and other animals. This ‘starvation-induced hyperactivity’ is a hallmark of foraging behavior32,38,40,50–53 as it enables animals to explore their environment and locate food sources. As major integrators of metabolic state information, IPCs are thought to be part of the neuronal circuitry that modulates foraging40. For example, insulin inhibits a subset of olfactory sensory neurons, such that decreased insulin release leads to an increase in sensitivity to food odors54. Our findings that IPC activity is directly modulated by starvation and feeding support the hypothesis that they are part of the neuronal networks modulating foraging. To further investigate the relationship between nutritional states, IPC activity, and foraging behavior, we used our ‘Universal Fly Observatory’ (UFO) setup to quantify locomotor activity55, in in particular starvation-induced hyperactivity (Figure 2A). In the UFO, flies can walk freely in a circular arena illuminated by infrared LEDs, and their walking activity can be quantified using semi-automated tracking software and analysis (based on the Caltech Fly Tracker56, see Methods, Figure 2A).

Walking activity is modulated by the nutritional state, OANs and IPCs.

A) Schematic showing the UFO setup. B) Average forward velocity (FV) of flies in different feeding states. Median and IQR are presented. p-values from Wilcoxon rank-sum test. C) Average FV of flies representing one replicate during optogenetic activation of OANs, in fed flies. Empty-GAL4 was used as control for all experiments (black). Red shading, optogenetic activation (300 s each). D) Average FV across all trials from two replicates for OAN activation: 300s before light onset, during stimulus, and after light offset. N = number of flies, n = number of activation trials. Thin lines represent individual trials, thick lines represent median of all trials. E) Average FV of all flies while activating OANs. Left: each stimulus trial (1-5) and Right: 300s after light offset (P1-P5). F) Average FV of flies representing one replicate during optogenetic activation of IPCs in fed flies. G) Average FV across all trials from two replicates for IPC activation in fed flies (detailed information as in D). H) Average FV of all flies while activating IPCs in fed flies (detailed information as in E). I) Average FV of flies representing one replicate during optogenetic activation of IPCs in starved flies. J) Average of FV across all trials from two replicates for IPC activation in starved flies (detailed information as in D). K) Average FV of all flies while activating IPCs in starved flies (detailed information as in K). L) and M) Average FV pooled across all activation trials (1-5) and post activation windows (P1-P5), respectively. Median and IQR are presented. p-values from Wilcoxon rank-sum test. Where no detailed p-value is stated, asterisks represent statistical significance. See also Table S1 and S2.

To test whether IPCs play a causal role in modulating starvation-induced hyperactivity, we first quantified the effects of nutritional manipulations on walking behavior. Fully fed flies, which have elevated IPC activity, displayed a low baseline walking activity with a median FV of 0.3 mm/s (Figure 2B). 24 h starved flies, in which IPC activity was significantly reduced, were more hyperactive and displayed a significantly increased median FV of 1.9 mm/s (Figure 2B). Refeeding flies on a pure-diet for 24 h after starvation, which led to a drastic increase in-IPC activity reduced the hyperactivity by about 50% (median 0.9 mm/s) compared to starved flies (Figure 2B). Hence, the hyperactivity was significantly reduced but not abolished, despite the fact that IPC activity was maximal in these flies (Figure 1G). One interpretation of this behavior is that flies are searching for protein sources. Flies refed on a full diet (SD) were even less active than flies fed ad libitum before undergoing starvation, and this difference was significant (Figure 2B). These results show that the locomotor activity was affected by the same dietary manipulations that had strong effects on IPC activity. However, IPC activity changes alone cannot fully explain the modulation of starvation induced hyperactivity, since high-glucose diets which drove the highest activity in IPCs were not sufficient to reduce locomotor activity back to baseline levels. This suggests that the modulation of starvation-induced hyperactivity is achieved by multiple modulatory systems.

Next, we investigated the temporal dynamics of the modulation of locomotor activity. Since nutritional state-dependent modulation of IPC activity occurred on a relatively slow timescale over several hours to days, we tested whether starvation-induced hyperactivity was modulated on a similar timescale. To this end, we quantified the walking activity of flies that were either fed ad libitum or starved for increasing durations, ranging from 30 minutes to 48 hours. As established earlier, flies that were fully fed or merely starved for 30 minutes had a low FV of 0.3 mm/s (Supplementary Figure S2A). However, their walking activity increased with increasing starvation duration up to 20 hours, when it reached a maximum at an FV of 3.6 mm/s (mean, Supplementary Figure 2A). The FV of flies starved for over 20 hours decreased again, probably due to a lack of energy reserves. We observed a final peak after 36 hours of starvation (Supplementary Figure S2A), which we interpret as a surge in starvation-induced hyperactivity before the metabolic state becomes critical. These results demonstrate that the starvation duration has a strong effect on locomotor activity, the modulation of which happens on a similar timescale as IPC modulation by changes in nutritional state.

Activation of octopaminergic neurons increases locomotor activity

Since changes in IPC activity alone were insufficient to explain starvation-induced hyperactivity, we next investigated which component of the behavioral changes were caused by IPCs and how strong their contribution was compared to other modulatory neurons. To accomplish this, we first activated octopaminergic neurons (OANs). Octopamine is the insect analogue to norepinephrine and is known to be released during flight, a metabolically demanding activity in Drosophila57,58, and to induce hyperactivity in starved flies40. Hence, we expected OANs to drive an increase in locomotor activity, rendering them a useful reference point. We used a tyrosine decarboxylase 2 (TDC2) driver line to acutely activate OANs via the optogenetic effector CsChrimson59 in the UFO while tracking the walking activity of flies. TDC2 is required for octopamine synthesis from tyramine60, therefore, TDC2-GAL4 labels all neurons producing octopamine and tyramine. In parallel, we used empty split-GAL461 flies crossed to UAS-CsChrimson as controls, which were simultaneously exposed to the same activation protocol in an adjacent UFO. This setup allowed us to account for activity changes due the red light used for optogenetic activation and other factors potentially affecting walking activity, such as small fluctuations in temperature, humidity, or differences in the circadian time. Each optogenetic activation experiment consisted of five activation cycles, in which we activated the neurons for 300 s, followed by an inter-stimulus-interval of 600 s, which we refer to as the ‘post activation’ window. Results reported here were analyzed by averaging two replicates per experiment.

We observed a strong, significant increase in FV from 0.23 mm/s (median) before OAN activation to 3.68 mm/s after five activation cycles, with a notable increase after each activation (Figure 2C and 2E, p < 0.001, see Supplementary Table S1 and S2). Notably, the final FV reached after OAN activation was comparable to that after 24 h of starvation (compare Figure 2E to starved controls in Figure 2B and I). This suggests that OAN activity can explain a large proportion of the starvation-induced hyperactivity. Interestingly, we observed a dramatic drop in FV at the onset of OAN activation (Figure 2D and 2E), during which the median FV dropped by 2 mm/s and approached 0 mm/s. This reduction in FV was due to recurring stopping behavior observed during acute OAN activation (Supplementary Figure S2B, Supplementary video S1). These behavioral effects of OAN activation have been previously described in adult flies as inhibition of food tracking via ventral unpaired median OANs and have been linked to suppression of long-distance foraging behavior62. Since these are not the only neurons we activate while activating tdc2 positive neurons, we speculate that the stopping phenotype could also result from concerted effects of octopamine and tyramine modulating muscle contractions, as previously described in Drosophila larvae63–65, or of OANs interfering with pattern generating networks in the ventral nerve cord (VNC) during long activation66. Overall, long-term optogenetic activation of OANs had a strong and significant effect on walking activity and mimicked the starvation-induced hyperactivity (Figure 2I and 2K, black and gray controls).

IPC activation slightly reduces starvation-induced hyperactivity

Having confirmed a strong effect of OANs on locomotor activity and the validity of our approach, we next examined the effects of IPCs on locomotor activity in response to changes in nutritional state. To achieve this, we optogenetically activated IPCs in freely walking flies, which were either fed ad libitum or wet-starved for 24 h. IPC activation in fed flies had only minor effects on the FV, indicating that IPCs do not drive starvation-induced hyperactivity or other aspects of foraging behavior (Figure 2F-H). The FV of most flies remained below 2 mm/s throughout the experiment (75th percentile, Figure 2H: see 1-5 and P1-P5) and was indistinguishable from the FV of controls during all but the first activation cycle (Figure 2F and 2H: p > 0.05). Accordingly, their FV did not differ during most activation windows (Median (1-5) = 0 - 0.5 mm/s, Figure 2H) or post-activation windows (Median (P1-P5) = 0 - 0.1 mm/s). One exception was the first activation cycle, in which the IPC-activated flies walked slightly but significantly faster than controls (Figure 2H, 1: p = 0.03). When averaging the activity of all flies across all activation windows, there was a small but statistically significant difference of 0.06 mm/s (Figure 2L, p= 0.01). In the post-activation windows, however, the FV of IPC activated and control flies were indistinguishable (Figure 2M). Overall, IPC activation had no notable effects on locomotor activity in fully fed flies.

Next, we tested the effect of IPCs on starvation-induced hyperactivity. Given that higher IPC activity levels are associated with the fed state, we expected that activating IPCs would reduce hyperactivity in starved flies. As previously observed (Figure 1B), starved flies were generally more active than their fed counterparts and walked with a median FV ≥ 2 mm/s (Figure 2I and K). The FV of flies during IPC activation remained significantly lower than the control flies over the course of all five activation cycles (Figure 2I-K, p < 0.05, Supplementary Table S1). This confirmed our hypothesis that IPC activation reduces starvation-induced hyperactivity. However, two observations were counterintuitive: First, the FV of IPC activated flies and control flies converged over time, and there was no significant difference in the control and IPC-activated flies during the last two post activation windows (Figure 2I, and 2K: P4, P5: p = 0.23 and p = 0.45, see also Supplementary Table S2). Hence, while activation of IPCs reduced hyperactivity, this effect did not persist after long-term activation. Second, we noted a decrease in the FV of IPC-activated starved flies even before the first optogentic stimulation (Figure 2I), possibly due to CsChrimson sensitivity to visible light. Overall, these results suggest that IPCs have a small but significant effect on behavior, in that they slightly reduced starvation-induced hyperactivity. We conclude that starvation-induced hyperactivity is affected by IPCs, but predominantly controlled by populations of modulatory neurons other than IPCs, for instance, OANs.

IPCs do not sense changes in extracellular glucose levels in vivo

Our experiments established that IPC activity is modulated by the nutritional state and that IPCs are sensitive to glucose ingestion, similar to pancreatic beta cells. Like beta cells, IPCs have been suggested to directly sense changes in extracellular glucose concentration in ex vivo studies23,43, where IPC activity increased in response to the application of saline containing a high-glucose concentration to the nervous system. Consequently, IPCs are thought to sense extracellular glucose increases cell-autonomously. To test this hypothesis in flies with fully intact sensory and central circuits, we recorded from IPCs in vivo, while perfusing extracellular saline containing high concentrations of glucose over the brain. After recording the IPC baseline activity in starved flies under perfusion of glucose-free saline (Figure 3A), we perfused the brain with saline containing 40 mM glucose (Figure 3A and 3B). Similar to our previous experiments, IPC spike rates were very low with a median of 0.05 Hz after starvation. Surprisingly, perfusing high concentrations of glucose over the brain did not increase IPC activity in starved flies (Figure 3B), even after 20 minutes of perfusion. In fed flies too, glucose perfusion had no effects on the IPC activity (Supplementary Figure S3A). Towards the end of each recording session, we injected current to establish that the IPCs were still excitable.

IPCs are not sensitive to glucose perfusion but DH44PINs are.

A) Schematic showing the experimental paradigm. IPC and DH44PIN baseline spike rates were recorded in glucose-free extracellular saline followed by recordings in glucose-rich saline. B) IPC spike rate and delta spike rate in glucose-free and glucose-rich extracellular saline, in starved flies. Spike rates were averaged within a five minute window. Delta spike rate was calculated by subtracting the baseline (Pre) from each trial for each IPC. Pre: 5-minute recording in glucose-free extracellular saline. Glucose-rich saline was allowed to perfuse for about eight minutes before analyzing IPC activity. Glu1 and Glu2: Two subsequent, 5-minute-long recordings in glucose-rich extracellular saline, starting eight minutes after onset of glucose perfusion. Each circle represents an individual IPC from a different fly, the thick line represents the grand mean of all recordings. p-values were calculated via Wilcoxon signed-rank test. C) Comparison of IPC baseline spike rate between starved, glucose-refed, and glucose-perfused flies highlighting the ‘incretin effect’. Median and IQR are indicated. D) Staining showing Drosophila brain with IPCs (magenta) and DH44Ns (green). GFP was enhanced with anti-GFP (green), brain neuropils were stained with anti-nc82 (cyan), and IPCs were labelled using a DILP2 antibody (magenta). E) Example membrane potential of a DH44PIN recorded in a fed (black) and a starved fly (orange). F) Baseline spike rate in fed (black) and starved flies (orange), each circle represents an individual DH44PIN. G) Comparison of the membrane potential of DH44PINs in fed and starved flies. Median and IQR are indicated. p-values calculated via Wilcoxon rank-sum test. H) Spike rate and delta spike rate of DH44PINs in glucose-free and glucose-rich extracellular saline, in starved flies. Thick line represents grand mean. p-value from Wilcoxon signed-rank test. I) Schematic showing the regulation of IPCs, DH44PINs and DH44Ns outside PI.

Our findings clearly show that IPCs do not directly sense glucose in their extracellular environment in vivo, but rather respond to glucose ingestion during feeding (Figure 3C). Hence, IPCs require input from the gut or other sensory pathways, such as taste or mechanosensation, to release insulin in response to glucose ingestion. Taken together, the increase in IPC activity following glucose ingestion but not glucose perfusion resembles the ‘incretin effect’ observed in mammals, where insulin secretion is significantly higher when glucose is ingested orally compared to isoglycemic intravenous delivery of glucose44,45. This effect is driven by gut-derived neuropeptides collectively referred to as ‘incretin hormones’ in mammals47. Recently, incretin-like hormones were also described in Drosophila43,66.

DH44PINs sense changes in extracellular glucose levels

We were surprised to find that IPCs were not responsive to glucose perfusion and needed to ensure this was not due to short-comings in our experimental protocol. To validate our approach, we recorded from DH44PINs. Dus and co-workers showed that DH44PINs are activated by D-glucose via calcium imaging experiments68. DH44PINs are anatomically similar to IPCs, and their cell bodies are located directly adjacent to those of IPCs in the PI, making them an ideal positive control for our experiments (Figure 3D). When recording from DH44PINs in vivo, we first observed that their spike amplitude fell within the same range as that of IPCs (Figure 3E, ∼40-60 mV). However, DH44PIN baseline spike rates were higher than those of IPCs, ranging from 0.5 Hz to 3 Hz (Figure 3F). Moreover, DH44PINs exhibited a firing pattern that was more burst-like (Figure 3E), with much shorter inter-spike-intervals (ISI) than IPCs (Supplementary Figure S3E). When comparing the ISI distribution between DH44PINs and IPCs, the fraction of ISIs below 500 ms was 60 % in DH44PINs and 19 % in IPCs. The cumulative probability distributions differed significantly (Supplementary Figure S3E, p << 0.001). Hence, the two neuron types displayed distinct activity patterns.

Next, we tested the effect of starvation on DH44PIN activity. The baseline spike rate was slightly lower in starved compared to fed flies, but, unlike in IPCs, this difference was not significant (Figure 3F). The membrane potential of DH44PINs was also not affected by starvation (Figure 3G). However, the cumulative probability distribution of DH44PIN ISIs differed significantly between starved and fed flies, and ISIs were shifted towards longer durations after starvation (p << 0.001, Supplementary Figure S3E). Hence, while the overall spike frequency was unaffected, DH44PINs spike patterns were more irregular and burst-like after feeding.

Finally, we tested whether DH44PINs are responsive to glucose perfusion. Following the exact same protocol we used for IPC recordings, we perfused the brain with saline containing 40 mM glucose, while patching from DH44PINs in vivo. The spike rate of DH44PINs significantly increased during perfusion with 40 mM glucose in starved flies, by about 0.5 Hz (Figure 3H). The ISI distribution was also significantly shifted towards shorter intervals during glucose perfusion across all flies and within each recording (Supplementary Figure S3F and G). Fed flies, on the other hand, did not exhibit a significant change in DH44PIN activity during glucose perfusion (Supplementary Figure S3D). This indicated that DH44PINs are indeed sensitive to glucose perfusion, and that the glucose sensitivity depends on the nutritional state, as previously reported68. Together, these experiments reveal two important points: First, DH44PINs are glucose sensing, whereas IPCs are not (Figure 3I). Second, the DH44PIN recordings confirm that that our negative result for IPC glucose sensitivity was not an experimental artifact.

DH44Ns outside the PI inhibit IPCs

The proximity of DH44PIN and IPC somata in the same cluster of the PI, their overlapping neurites in the brain (Figure 3D), and the fact that both populations are implicated in metabolic homeostasis prompted us to determine whether these neurons directly interact with each other. To investigate potential functional connectivity between DH44Ns and IPCs, we combined the UAS-GAL4 and LexA-LexAop systems to simultaneously express CsChrimson in all DH44Ns and GFP in IPCs (Figure 4A, see also Supplementary Figure S4A)27. This approach enabled us to optogenetically activate DH44Ns while recording from IPCs via patch-clamp.

DH44Ns outside the PI inhibit IPCs and drive increases in locomotor activity

A) Immunolabelling showing DH44 expression in the brain and the VNC from broad DH44-GAL4 driver line. GFP was enhanced with anti-GFP (green), brain neuropils were stained with anti-nc82 (magenta) B) Example recording of an IPC during optogenetic activation of the DH44Ns (red shading). Upper panel shows individual trials, lower panel shows ten trials overlapped and the median of all trials (brown trace). C) Upper panel shows the spike density of individual IPCs across 10 DH44N activations. Lower panel shows the baseline-subtracted, median filtered Vm traces for each IPC. Thick lines represent the grand mean. D) Effect of DH44N activation on IPCs. Delta Vm is plotted by calculating the median baseline subtracted Vm from C) 500 ms before (Pre) and 200 ms after DH44N activation (Post). Each circle represents one IPC recording. p-values from Wilcoxon signed-rank test. E) Immunolabelling showing GFP expression in the brain and the VNC in the sparse DH44PI-GAL4 driver line. F) Example recording of an IPC during optogenetic activation of DH44 neurons using a sparse line which labels only DH44PINs. Plot details as in B). G) Spike density and baseline subtracted medians of individual IPCs while activating DH44PI line. Plot details as in C). H) Pre and post delta Vm of IPCs before and after optogenetic activation of the DH44PI line. Plot details as in in D). I) Average FV of 20 flies during optogenetic activation of the DH44Ns using the broad driver line (A). Empty-GAL4 was used as control for all experiments (black). Red shading shows 300s activation windows. J) Average FV across all DH44N activation trials based on two independent replications of the experiment in I. K) Average FV of all flies during each stimulus trial (1-5) and post-stimulus trial (300 s window immediately after activation seized, P1-P5). Circles and bars show median and IQR, respectively. Asterisks represent a significant difference according to a Wilcoxon rank-sum test. L-N) Behavioral effects of optogenetic DH44PIN activation (see E). Plot details as in I, J and K, respectively. See also Table S3 and S4.

Optogenetic activation of DH44Ns drove an immediate, strong hyperpolarization of IPCs, which lasted for over one second after stimulus cessation and reduced IPC spiking repeatedly across trials (Figure 4B). This effect was robust across the population of IPCs and flies (Figure 4C). On average, the IPC membrane potential dropped by -5 mV below the baseline during DH44N activation and stayed hyperpolarized until about 1s after activation (Figure 4C). This fast and strong response suggests that DH44Ns could be directly presynaptic to IPCs. To quantify this effect, we calculated the median Vm for each IPC recording in a 500 ms window before the stimulus onset (pre, time windows indicated by grey and colored boxes in Figure 4C) and 200 ms after cessation of the stimulus (post, Figure 4D). All IPCs recorded were strongly hyperpolarized by DH44N activation (Figure 4D, p = 0.002). Hence, DH44Ns strongly and significantly inhibited IPCs. Notably, we also recorded from one IPC, which was excited during DH44N activation and inhibited immediately after cessation of activation (Supplementary Figure S4C), indicating heterogeneity in the IPC population30. Overall, our results indicate that acute activation of DH44Ns inhibits IPC activity, hence regulating insulin release. This strong inhibition is surprising, since our previous results implicated both, the DH44PINs and IPCs, in glucose sensing and metabolic control. Hence, our assumption had been that these modulatory neurons, would display reciprocal excitation rather than inhibition. However, the driver line we used for DH44N activation was relatively broad, and any of the neurons labelled could be responsible for the IPC inhibition. For example, the line labels a subset of DH44Ns ascending from the VNC. A pair of these DH44Ns co-express the neuropeptide Leucokinin (LK)69, and LK expressing neurons inhibit most IPCs30.

Therefore, we identified a driver line which only labels the DH44Ns in the PI70 (DH44PI–GAL4) and repeated the functional connectivity experiments using this more specific driver line (Figure 4E, see also Supplementary Figure S4B). Activating exclusively DH44PINs had no effect on the IPC membrane potential or spike rate (Figure 4F and, 4G: grand mean remained 0 mV on average). This was consistent across the population of IPCs and flies (Figure 4H, p = 0.641). These results confirm that the strong inhibition of IPCs observed when activating the broad DH44N line is not driven by DH44PINs. Therefore, DH44PINs and IPCs both sense changes in the nutritional state-but act independently and in parallel to signal changes in nutritional state (Figure 3I).

Activation of DH44Ns modulates locomotor activity

Since IPC manipulation had relatively small effects on locomotor activity, other pathways need to be responsible for the strong changes in locomotor activity observed during starvation (Figure 2B). Given their sensitivity for glucose and connectivity to IPCs, DH44Ns were strong candidates. DH44Ns have previously been shown to promote feeding in starved animals68. In particular, we hypothesized that the broad DH44N line could have opposite effects to IPCs since a subset of them inhibited IPCs. We therefore tested whether DH44Ns affect locomotor activity by following the same approach as for IPC activation in freely walking flies (Figure 2I-K). Activation of the broad DH44N line resulted in a dramatic increase in FV from a median of 1.8 mm/s to 6 mm/s immediately after light onset (Figure 4I, 4J). This increase was followed by an abrupt halt lasting until the end of the DH44N activation (Figure 4I-K). During this time window of reduced locomotor activity, we observed a strong increase in proboscis extension in 6/20 flies for which we quantified this (Supplementary Figure S4D). The locomotor arrest was released soon after cessation of DH44N activation, upon which locomotor activity increased until the onset of the next stimulation period (Figure 4I and 4K). The FV in the post-activation window increased over subsequent activation cycles and was significantly higher than in controls throughout the experiment (Figure 4I, 4K). The FV of DH44N activated flies increased from a median of 0.3 mm/s to 3 mm/s over five activation cycles, suggesting that long term activation of DH44Ns leads to an increase in locomotion comparable to starvation-induced hyperactivity.

In addition, we tested the driver line labeling only DH44PINs. Although we did observe an overall increase in FV of the experimental group, we neither observed an activity peak upon activation onset, nor locomotor arrest during activation (Figure 4L-N). Hence, the strong activation phenotypes were driven by DH44Ns outside the PI. We conclude that the activation of DH44Ns, especially the broad DH44N line, had a pronounced and sustained effect on foraging behavior. The same neurons strongly and significantly inhibited IPCs, suggesting that feeding-promoting and satiety-inducing pathways form inhibitory connections in the CNS as a part of the neuronal mechanisms to ensure reciprocal inhibition of satiety and hunger states.

Discussion

IPC activity is modulated on short and long timescales by the nutritional state, the behavioral state, and aging processes

To advance our understanding of the neuronal underpinnings of metabolic homeostasis, we analyzed the nutritional state-dependent modulation of IPCs in Drosophila. Our in vivo patch clamp recordings revealed that IPC activity is diminished during starvation and rebounds after glucose feeding. This is consistent with previous studies demonstrating an accumulation of DILPs26,71–75 and a reduction of dilp transcript levels72,74,75 in IPCs of adult and larval fly brains under starvation and underlines that electrophysiological IPC activity and DILP release are correlated. However, two aspects were surprising about our results. First, the rebound after glucose feeding was relatively slow, and IPC activity built up to baseline levels after over six hours of refeeding. These slow temporal dynamics are a stark contrast to the behavioral state-dependent modulation of IPC activity, which occurs on a timescale of milliseconds27. This suggests that IPCs integrate changes in internal states over different timescales and potentially affect metabolism and neuronal networks on both short and long timescales. This idea was further supported by the fact that IPC activity was heavily diminished in flies older than ten days (Supplementary Fig 1B). Reduced insulin signalling has previously been linked to increased life-span in different species, including yeast, worms, flies, and rodents1. Our findings demonstrate that IPC activity is significantly reduced both in case of starvation and aging, indicating that insulin release from IPCs is suppressed in these conditions. This suppression of IPC activity could be a mechanism to increase survival under physiologically unfavourable conditions74. DILP6 mRNA in the fat body is upregulated during starvation, which has been shown to decrease systemic insulin signaling by the suppression of insulin release from IPCs. This is linked to an increased life-span in Drosophila74. The second surprising result was that the reduced IPC activity after starvation failed to rebound upon refeeding with high protein and standard diets (Figure 1H). Our standard diet contains glucose at very low concentrations (upper bound of 30 mM, see Methods for details), which explains the lack of an IPC activity rebound. This suggests that starvation had a long-lasting effect on IPC activity, which could specifically be restored by glucose feeding. In addition, insulin signaling might be impaired in flies if they only feed on protein. Similar observations were made in mammals, where prolonged high protein diets lead to insulin resistance and type II diabetes76,77. Furthermore, dilp2 mutants show an increased lifespan78 which likely results from elevated glycogen phosphorylase activity79, an enzyme involved in regulating glycogen storage. This suggests that suppressed IPC activity in aged flies could mediate glucose homeostasis by increasing glycogen catabolism. Taken together, the reduced IPC activity we observed in starved and aged flies may be crucial for metabolic adjustments required for survival.

The nutritional state-dependent modulation of IPCs supports the presence of an incretin-like effect in Drosophila

Mammalian pancreatic β-cells sense glucose levels in the blood and release insulin accordingly80,81. However, this process is much more complex than initially assumed. One key observation is that the ingestion of glucose drives much higher release of insulin compared to the isoglycemic intravenous perfusion of glucose. This difference in insulin release is partially driven by incretin hormones from the gut, for example glucagon-like peptide-1 (GLP-1) and glucose-dependent insulinotropic peptide (GIP)82–84, which are released after the ingestion of food. Hence, insulin release is modulated by enteroendocrine signals. Drosophila IPCs are functionally analogous to mammalian β-cells and presumably sense glucose in the circulating hemolymph to release insulin in response to increases in glucose concentration. Previous ex vivo studies suggested that IPCs, like pancreatic β-cells, sense glucose cell-autonomously23,43. Consistent with this, we observed an increase in IPC activity after the ingestion of glucose (Fig 2B). However, IPC activity did not increase during the perfusion of glucose directly over the brain. Importantly, the fly preparations were kept alive for several hours allowing the glucose-rich saline to enter circulation and reach all body parts. This suggests that, similar to mammals, IPC activity and hence, insulin release, is not simply modulated by hemolymph glucose concentration in Drosophila. Several pathways could contribute to this. First, although an incretin effect has not been reported in flies before, recent studies provided evidence of incretin-like hormones such as the midgut-derived neuropeptide F in flies85,86, suggesting that a similar signaling pathway could underlie the incretin effect across species. Second, a subset of IPCs was shown to express the mechano-sensitive channel piezo and innervate the crop87, suggesting that an increase in crop volume during feeding could contribute to the increase in IPC activity after feeding. However, the fact that we observed strong responses of IPCs to glucose intake but not the ingestion of regular food during refeeding suggests this is not the core mechanism underlying the incretin-like effect we observed. Third, taste pathways converge onto IPCs88 and they could increase IPC activity after the ingestion of glucose via serotonergic inputs from taste responsive cells in the gnathal ganglion. Here, however, we would expect faster responses on the scale of seconds, which are unlikely to shift IPC activity over several hours. Hence, gut-secreted neuropeptides are the strongest candidates to mediate the incretin effect in Drosophila. These results signify a causal link between glucose homeostasis in the gut and the brain, and pave the way for further studies investigating the evolutionary conservation of the incretin effect across mammals and insects.

DH44PINs and IPCs have distinct, yet interrelated roles in nutrient sensing

Our in vivo recordings revealed that IPCs do not sense glucose cell-autonomously, and neither do neurons in the CNS that are directly connected to IPCs – otherwise we would have expected an increase in IPC activity during glucose perfusion. In contrast, DH44PINs were sensitive to changes in the extracellular glucose concentration. This could either be driven by cell-autonomous glucose sensitivity or presynaptic glucose sensing neurons. Moreover, DH44PINs did not show an increase in activity in fed flies, indicating a lack of sensitivity to glucose ingestion. IPCs responded to glucose ingestion on a timescale of hours, whereas DH44PINs immediately responded to increases in extracellular glucose concentrations on a timescale of seconds. These results suggest that DH44PINs and IPCs play distinct roles in glucose sensing, which might complement each other. Our initial functional connectivity experiments suggested that the glucose-sensing DH44PINs might inhibit IPCs directly. However, refined experiments using a sparse driver line revealed that IPCs are inhibited by DH44Ns outside the PI. If these DH44Ns, like DH44PINs, are glucose sensing they would suppress IPC activity in the presence of high glucose concentrations, which would be counteractive. Therefore, even though this remains to be determined, it is likely that these non-PI DH44Ns are activated by signals other than glucose concentrations. Regardless, our re-

sure to visible light during fly handling. However, with subsequent activations, the difference between IPC-activated and control flies was reduced, probably because flies were increasingly food and water-deprived after an hour of activation experiments. In particular the latter could potentially override the reduction in starvation-induced hyperactivity induced by optogenetic IPC activation. Overall, our experiments revealed that IPC activation and resulting insulin release do not lead to a sustained satiety response. This aligns with physiological expectations, as elevated insulin levels exacerbate the glucose deficit in starved flies, which could ultimately lead to increases in hyperactivity via this secondary effect.

sults shows that the glucose sensing DH44PINs and IPCs act independently of each other. This is coherent with what we established so far: brain DH44PINs increase their activity upon glucose sensing.

In line with previous studies68,89, we found that DH44PINs were only activated by glucose in starved flies. Interestingly, this was not because the DH44PIN spike rate was lower after starvation. Instead, DH44PINs exhibited similar spike rates in fed and starved flies, but the activity increase in response to glucose perfusion only occurred under starvation. DH44PINs also displayed a level of modulation beyond simple increases and decreases in spike frequencies. For example, the ISI distributions of DH44PINs were shifted to significantly longer intervals during starvation. This may contribute to the increased glucose sensitivity in starved flies. Moreover, shorter ISIs in fed flies and the more bursty activity pattern could lead to an increase in DH44 peptide release despite similar overall spike frequencies. Apart from this, DH44PINs themselves could also become more sensitive to glucose in starved animals. This would suggest that DH44PINs sense changes in glucose concentration, rather than the absolute concentration of glucose, in the brain or hemolymph.

Nutritional state-dependent modulation of behaviour

Changes in the activity of modulatory neurons affect information processing in neuronal circuits and ultimately behavior22,54,90–92. This includes starvation-induced hyperactivity and other aspects of foraging in Drosophila. We first tested the causal role of one of the known neuromodulators involved in driving starvation-induced hyperactivity – octopamine. As expected, OAN activation increased locomotor activity, which we interpret as a proxy for foraging based on previous publications38,93. The activity levels reached by long-term OAN activation were comparable to the effects of long-term starvation. These results underscore that octopamine plays a major role in driving starvation-induced hyperactivity. Having established the method and approach, we next investigated whether IPCs modulate locomotor activity in flies. This was prompted by the observation that the same parameters that affected IPC activity, i.e., starvation and different diets, also affected the baseline walking activity. However, IPCs only had a minor effect on locomotion. This was despite stimulating IPCs using long, high-intensity optogenetic activation. The strongest effect we observed was a reduction in starvation-induced hyperactivity upon IPC activation, which we interpret as a satiety-like effect due to increased IPC activity. This effect was rather strong in the beginning of the experiment, even before the first activation of IPCs, and was robust across independent repetitions of the same experiment. This effect was likely enhanced by an increased IPC depolarization via CsChrimson activation because of expo

Our results also suggest a key role of DH44Ns in modulating aspects of foraging behavior. There is evidence that DH44 release in the brain modulates rest:activity rhythms94 via downstream targets in the gnathal ganglion95, which houses circuits involved in the modulation of feeding and locomotion34,95–100. Our optogenetic approach provides substantial evidence that DH44Ns drive increased locomotor activity in adult flies, plausibly to enable exploration of the environment for locating food sources. First, the increase in locomotion phenocopies starvation-induced hyperactivity. Second, our experiments additionally showed that DH44Ns outside the PI drive proboscis extensions during stopping, which are one hallmark of feeding behavior. In line with this, a previous study has shown that activation of DH44 receptor 1 neurons, the target neurons of the DH44 neuropeptide, induces proboscis-extension68. Both observations suggest that DH44Ns are a component of feeding pathways and mediate nutrient uptake in order to maintain nutritional homeostasis.

Our functional connectivity experiments revealed that these DH44Ns inhibit IPC activity. Initially, promoting feeding and inhibiting insulin release at the same time seems counter-regulatory. However, there are two possible explanations: First, these two effects could be driven by different sub-populations of DH44Ns labelled by the line we used (Fig 4A and E). Second, the IPC inhibition could be part of a combination of physiological effects signaling a hunger state and contribute to the increased foraging activity driven by DH44Ns. Notably, blocking insulin signalling has been reported to mimic starvation in fed flies54. Contrary to DH44Ns outside the PI, activation of the glucose-sensing DH44PINs did not drive proboscis extension, suggesting that DH44PINs are not involved in feeding initiation, but primarily function in post feeding regulation of metabolic homeostasis. This is in line with our electrophysiological results showing that DH44PIN spike frequency is not affected by starvation, and only starts to increase once glucose levels change in the hemolymph. Furthermore, DH44PIN activation increased the FV but did not lead to stopping, suggesting that DH44PINs contribute to starvation-induced hyperactivity, whereas other DH44Ns drive stopping.

In conclusion, our study sheds light onto the intricate connections between neuroendocrine signaling, nutrient sensing, and behavior in Drosophila. Using electrophysiological approaches, we unraveled the complex activity dynamics of IPCs and DH44PINs and identified differences in their activity patterns during various nutritional states. The discovery of an ‘incretin-like’ effect in flies suggests that important aspects of gutbrain signalling are conserved across vertebrate and invertebrate species. Hence, our findings not only contribute to the increasing body of research on insulin signaling, but also pave the way for future research projects analyzing the intricacies of Drosophila neuroendocrine networks governing nutrient sensing.

eLife Assessment

This study reports on the in vivo dynamics of insulin-producing cells (IPCs) in Drosophila. IPC activity is shown to be modulated by the nutritional state and age of the animal, with convincing evidence for an incretin-like effect. These important findings establish IPCs in Drosophila as a system to study circuits governing behaviors related to the internal state in competition with the feeding state, and will be of interest to both neuroscientists and cell biologists.

Reviewer #1 (Public Review):

Summary:

This study presents useful insights into the in vivo dynamics of insulin-producing cells (IPCs), key cells regulating energy homeostasis across the animal kingdom. The authors provide compelling evidence using adult Drosophila melanogaster that IPCs, unlike neighboring DH44 cells, do not respond to glucose directly, but that glucose can indirectly regulate IPC activity after ingestion supporting an incretin-like mechanism in flies, similar to mammals. The authors link the decreased activity of IPCs to hyperactivity observed in starved flies, a locomotive behavior aimed at increasing food search.

Furthermore, there is supporting evidence in the paper that IPCs receive inhibitory inputs from Dh44 neurons, which are linked to increased locomotor activity. However, although the electrophysiological data underlying the dynamics of IPCs in vivo is compelling, the link between IPCs and other potential elements of the circuitry (e.g. octopaminergic neurons) regulating locomotive behaviors is not clear and would benefit from more rigorous approaches.

This paper is of interest to cell biologists and electrophysiologists, and in particular to scientists aiming to understand circuit dynamics pertaining to internal state-linked behaviors competing with the feeding state, shown here to be primarily controlled by the IPCs.

Strengths:

(1) By using whole-cell patch clamp recording, the authors convincingly showed the activity pattern of IPCs and neighboring DH44 neurons under different feeding states.

(2) The paper provides compelling evidence that IPCs are not directly and acutely activated by glucose, but rather through a post-ingestive incretin-like mechanism. In addition, the authors show that Dh44 neurons located adjacent to the IPCs respond to bath application of glucose contrary to the IPCs.

(3) The paper provides useful data on the firing pattern of 2 key cell populations regulating food-related brain function and behavior, IPCs and Dh44 neurons, results which are useful to understand their in vivo function.

Weaknesses:

(1) The term nutritional state generally refers to the nutrients which are beneficial to the animal. In Figure 1, the authors showed that IPCs respond to glucose but not proteins. To validate the term nutritional state the authors could test the effect of a non-nutritive sugar (e.g. D-arabinose or L-Glucose) on the post-ingestive physiological responses of the IPCs.

(2) It is difficult to grasp the main message from the figures in the result section as some figures have several results subsections referring to different points the authors want to make. The key results of a figure will be easier to understand if they are summarized in one section of the results. Alternatively, a figure can be split into 2 figures if there are several key messages in those figures, e.g. Figures 2 and 3.

(3) The prime investigation of the paper is about the physiological response and locomotive behavioral readout linked to IPCs. The authors do not show a link between OANs and IPCs in terms of functional or behavioral readouts. In Figure 2 the authors first start with stating a link between OAN neurons and locomotion changes resulting from internal feeding states. The flow of the paper would be better if the authors focused on the effect of optogenetic activation of IPCs under different feeding states and their impact on fly locomotion. If the experiments done on optogenetic activation of OANs were to validate the experimental approach the data on OAN neurons is better suited for the supplement without the need of a subsection in the result section on the OANs.

(4) Figure 2F shows that optogenetic activation of IPCs in fed flies does not influence their locomotor output. In the text, the conclusion linked to Figure 2F-H states that IPC activation reduces starvation-induced hyperactivity which is a statement more suited to Figure 2I-K.

(5) The authors show activation of Dh44 neurons leads to hyperpolarisation of the IPCs. What is the functional link between non-PI Dh44 neurons and the IPCs? Do IPCs express DH44R or is DH44 required for this effect on IPCs? Investigating a potential synaptic or peptidergic link between DH44 neurons and IPCs and its effect on behavior would benefit the paper, as it is so far not well connected.

Reviewer #2 (Public Review):

Summary:

In this study, Bisen et al. characterized the state-dependency of insulin-producing cells in the brain of *Drosophila melanogaster*. They successfully established that IPC activity is modulated by the nutritional state and age of the animal. Interestingly, they demonstrate that IPCs respond to the ingestion of glucose, rather than to perfusion with it, an observation reminiscent of the incretin effect in mammals. The study is well conducted and presented and the experimental data convincingly support the claims made.

Strengths:

The study makes great use of the tools available in *Drosophila* research, demonstrating the effect that starvation and subsequent refeeding have on the physiological activity of IPCs as well as on the behavior of flies to then establish causal links by making use of optogenetic tools.

It is particularly nice to see how the authors put their findings in context to published research and use for example TDC2 neuron activation or DH44 activity to establish baselines to relate their data to.

Weaknesses:

I find the inability of SD to rescue the IPC starvation effect in Figure 1G&H surprising, given that the fully fed flies were raised and kept on that exact diet. Did the authors try to refeed flies with SD for longer than 24 hours? I understand that at some point the age effect would also kick in and counteract potential IPC activity rescue. I think the manuscript would benefit if the authors could indicate the exact age of the SD refed flies and expand a bit on the discussion of that point.

The incretin-like effect is exciting and it will be interesting in the future to find out what might be the signal mediating this effect. It is interesting that IPCs in explants seem to be responsive to glucose. I think it would help if the authors could briefly discuss possible sources for the different findings between these in fact very different preparations. Could the the absence of the inhibitory DH44 feedback in the *ex-vivo* recordings for example play a role?

The incretin-like effect the authors observed seems to start only after 5h which seems longer than in mammals where, as far as I know, insulin peaks around 1h. Do the authors have ideas on how this timescale relates to ingestion and glucose dynamics in flies?

The authors mention "a decrease in the FV of IPC-activated starved flies even before the first optogenetic stimulation (Figure 2I),". Could this be addressed by running an experiment in darkness, only using the IR illumination of their behavioral assay?

The authors show an inhibitory effect of DH44 neuron activation on IPC activity. They further demonstrate that DH44PI neurons are not the ones driving this and thus conclude that "...IPCs are inhibited by DH44Ns outside the PI.". As the authors mentioned the broad expression of the DH44-Gal4 line, can they be sure that the cells labeled outside the PI are actually DH44+? If so they should state this more clearly, if not they should adapt the discussion accordingly.

Reviewer #3 (Public Review):

Although insulin release is essential in the control of metabolism, adjusted to nutritional state, and plays major roles in normal brain function as well as in aging and disease, our knowledge about the activity of insulin-producing (and releasing) cells (IPCs) in vivo is limited.

In this technically demanding study, IPC activity is studied in the Drosophila model system by fine in vivo patch clamp recordings with parallel behavioral analyses and optogenetic manipulation.

The data indicate that IPC activity is increased with a slow time course after feeding a high-glucose diet. By contrast, IPC activity is not directly affected by increasing blood glucose levels. This is reminiscent of the incretin effect known from vertebrates and points to a conserved mechanism in insulin production and release upon sugar feeding.

Moreover, the data confirm earlier studies that nutritional state strongly affects locomotion. Surprisingly, IPC activity makes only a negligible contribution to this. Instead, other modulatory neurons that are directly sensitive to blood glucose levels strongly affect modulation. Together, these data indicate a network of multiple parallel and interacting neuronal layers to orchestrate the physiological, metabolic, and behavioral responses to nutritional state. Together with the data from a previous study, this work sets the stage to dissect the architecture and function of this network.

Strengths:

State-of-the-art current clamp in situ patch clamp recordings in behaving animals are a demanding but powerful method to provide novel insight into the interplay of nutritional state, IPC activity, and locomotion. The patch clamp recordings and the parallel behavioral analyses are of high quality, as are the optogenetic manipulations. The data showing that starvation silences IPC activity in young flies (younger than 1 week) are compelling. The evidence for the claim that locomotor activity is not increased upon IPC activity but upon the activity of other blood glucose-sensitive modulatory neurons (Dh44) is strong. The study provides a great system to experimentally dissect the interplay of insulin production and release with metabolism, physiology, and behavior.

Weaknesses:

Neither the mechanisms underlying the incretin effect, nor the network to orchestrate physiological, metabolic, and behavioral responses to nutritional state have been fully uncovered. Without additional controls, some of the conclusions would require significant downtoning. Controls are required to exclude the possibility that IPCs sense other blood sugars than glucose. The claim that IPC activity is controlled by the nutritional state would require that starvation-induced IPC silencing in young animals can be recovered by feeding a normal diet. At current firing in starvation, silenced IPCs can only be induced by feeding a high-glucose diet that lacks other important ingredients and reduces vitality. Therefore, feasible controls are needed to exclude that diet-induced increases in IPC firing rate are caused by stress rather than nutritional changes in normal ranges. The finding that refeeding starved flies with a standard diet had no effect on IPC activity but a strong effect on the locomotor activity of starved flies contradicts the statement that locomotor activity is affected by the same dietary manipulations that affect IPC activity. The compelling finding that starvation induces IPC firing would benefit from determining the time course of the effect. The finding that IPCs are not active in fed animals older than 1 week is surprising and should be further validated.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This study presents useful insights into the in vivo dynamics of insulin-producing cells (IPCs), key cells regulating energy homeostasis across the animal kingdom. The authors provide compelling evidence using adult Drosophila melanogaster that IPCs, unlike neighboring DH44 cells, do not respond to glucose directly, but that glucose can indirectly regulate IPC activity after ingestion supporting an incretin-like mechanism in flies, similar to mammals. The authors link the decreased activity of IPCs to hyperactivity observed in starved flies, a locomotive behavior aimed at increasing food search.

Furthermore, there is supporting evidence in the paper that IPCs receive inhibitory inputs from Dh44 neurons, which are linked to increased locomotor activity. However, although the electrophysiological data underlying the dynamics of IPCs in vivo is compelling, the link between IPCs and other potential elements of the circuitry (e.g. octopaminergic neurons) regulating locomotive behaviors is not clear and would benefit from more rigorous approaches.

This paper is of interest to cell biologists and electrophysiologists, and in particular to scientists aiming to understand circuit dynamics pertaining to internal state-linked behaviors competing with the feeding state, shown here to be primarily controlled by the IPCs.

Strengths:

(1) By using whole-cell patch clamp recording, the authors convincingly showed the activity pattern of IPCs and neighboring DH44 neurons under different feeding states.

(2) The paper provides compelling evidence that IPCs are not directly and acutely activated by glucose, but rather through a post-ingestive incretin-like mechanism. In addition, the authors show that Dh44 neurons located adjacent to the IPCs respond to bath application of glucose contrary to the IPCs.

(3) The paper provides useful data on the firing pattern of 2 key cell populations regulating foodrelated brain function and behavior, IPCs and Dh44 neurons, results which are useful to understand their in vivo function.

Weaknesses:

(1) The term nutritional state generally refers to the nutrients which are beneficial to the animal. In Figure 1, the authors showed that IPCs respond to glucose but not proteins. To validate the term nutritional state the authors could test the effect of a non-nutritive sugar (e.g. D-arabinose or L-Glucose) on the post-ingestive physiological responses of the IPCs.

We thank the referee for this insightful comment. Following their suggestion, we included two new experimental data sets, which we added to Figure 1: We show that IPCs do not respond to the non-nutritive sugar D-arabinose (Figure 1H). In order to further expand this data set and our conclusions, we additionally show that IPCs do respond to fructose – a second nutritive sugar in addition to glucose (Figure 1H). Together, these data sets permit the conclusion that IPCs are sensitive to the ingestion of nutritive sugars, and do not respond to ingestion of nonnutritive sugars or high protein diets. Thus, we validate the term nutritional state.

(2) It is difficult to grasp the main message from the figures in the result section as some figures have several results subsections referring to different points the authors want to make. The key results of a figure will be easier to understand if they are summarized in one section of the results. Alternatively, a figure can be split into 2 figures if there are several key messages in those figures, e.g. Figures 2 and 3.

We appreciate this suggestion and have made several changes to our manuscript to add more clarity. Among other things, we have changed the order of data presentation in Figure 2, as suggested by the referee below, where we now start with the IPC activation data rather than the OAN activation. We also swapped the order of data presentation and split Figure S1 into Figures S1 & S2. Moreover, we re-arranged the panel order in supplementary figure S4. This significantly improved the flow of the results section. Since the figures the referee refers to contain comparative data, for example between diets (Figure 1) or neuron types (Figure 2), we prefer to keep these data sets together. However, we have carefully revised the results section to more clearly relate our statements to individual figure panels.

(3) The prime investigation of the paper is about the physiological response and locomotive behavioral readout linked to IPCs. The authors do not show a link between OANs and IPCs in terms of functional or behavioral readouts. In Figure 2 the authors first start with stating a link between OAN neurons and locomotion changes resulting from internal feeding states. The flow of the paper would be better if the authors focused on the effect of optogenetic activation of IPCs under different feeding states and their impact on fly locomotion. If the experiments done on optogenetic activation of OANs were to validate the experimental approach the data on OAN neurons is better suited for the supplement without the need of a subsection in the result section on the OANs.

We agree with the reviewer’s suggestion and switched the order of the figure panels and text to aid the flow of the manuscript. We now show and discuss the IPC activation data first (Figure 2C-H) and OAN activation afterwards (Figure 2I-K). We did keep the OAN data in the main document, though, since that facilitates comparisons between the small effects of IPC activation and the large, well-established effects of OAN activation.

(4) Figure 2F shows that optogenetic activation of IPCs in fed flies does not influence their locomotor output. In the text, the conclusion linked to Figure 2F-H states that IPC activation reduces starvation-induced hyperactivity which is a statement more suited to Figure 2I-K.

We edited the text accordingly.

(5) The authors show activation of Dh44 neurons leads to hyperpolarisation of the IPCs. What is the functional link between non-PI Dh44 neurons and the IPCs? Do IPCs express DH44R or is DH44 required for this effect on IPCs? Investigating a potential synaptic or peptidergic link between DH44 neurons and IPCs and its effect on behavior would benefit the paper, as it is so far not well connected.

Although we have not performed any experiments dedicated to investigating the functional link between DH44Ns outside the PI and the IPCs in this study, there are two lines of evidence supporting that this connection is relatively direct. First, IPCs do express DH44R1 & R2, as we show in a parallel study in eLife (Held M, et al. ‘Aminergic and peptidergic modulation of Insulin-Producing Cells in Drosophila’. eLife. 2024;13. doi:10.7554/ELIFE.99548.1). Second, we performed functional connectivity experiments using a Leucokinin (LK) driver line in that paper. This driver line labels two pairs of non-PI DH44Ns in the VNC, which are DH44 and LK positive (Zandawala et al 2018). Activating that line leads to inhibition of IPCs, similar to the effect we observed here for DH44N activation. These two lines of evidence suggest that there could be a direct peptidergic connection between DH44+ neurons and IPCs. We have added a paragraph mentioning these experiments to our discussion:

‘Notably, the DH44PINs express the DH44 peptide, as confirmed by anti-DH44 stainings(100). This also applies to a large fraction of neurons labelled in the broad DH44 driver line(100). However, a subset of neurons labelled in the broad line did not exhibit DH44 immunoreactivity(100), and might therefore not actually express the DH44 peptide. Hence, the inhibition of IPCs could be driven by neurons in the DH44 driver line that do not express DH44. A strong candidate for the inhibition are LK and DH44-positive neurons, which are labelled by the broad line(76). In a parallel study, we showed that LK-expressing neurons strongly inhibit IPCs(30), similar to the broad DH44 line used here. Furthermore, evidence from single-nucleus transcriptomic analysis shows that IPCs express DH44-R1 and DH44-R2 receptors(30). Therefore, it is possible that DH44Ns communicate with IPCs through a direct peptidergic connection. Notably, the inhibitory effect of non-PI DH44Ns on IPCs was very strong and fast, suggesting that a connection via classical synapses is more likely. Regardless, our results show that the glucose sensing DH44PINs and IPCs act independently of each other.’

Reviewer #2 (Public Review):

Summary:

In this study, Bisen et al. characterized the state-dependency of insulin-producing cells in the brain of *Drosophila melanogaster*. They successfully established that IPC activity is modulated by the nutritional state and age of the animal. Interestingly, they demonstrate that IPCs respond to the ingestion of glucose, rather than to perfusion with it, an observation reminiscent of the incretin effect in mammals. The study is well conducted and presented and the experimental data convincingly support the claims made.

Strengths:

The study makes great use of the tools available in *Drosophila* research, demonstrating the effect that starvation and subsequent refeeding have on the physiological activity of IPCs as well as on the behavior of flies to then establish causal links by making use of optogenetic tools.

It is particularly nice to see how the authors put their findings in context to published research and use for example TDC2 neuron activation or DH44 activity to establish baselines to relate their data to.

Weaknesses:

I find the inability of SD to rescue the IPC starvation effect in Figure 1G&H surprising, given that the fully fed flies were raised and kept on that exact diet. Did the authors try to refeed flies with SD for longer than 24 hours? I understand that at some point the age effect would also kick in and counteract potential IPC activity rescue. I think the manuscript would benefit if the authors could indicate the exact age of the SD refed flies and expand a bit on the discussion of that point.

We have expanded the first paragraph of our discussion to tackle these questions, in particular the potential effect of aging, as suggested by the referee. We now also indicate the exact age of the flies. Moreover, we have conducted additional experiments in which we added either glucose or arabinose to our standard diet (Figure 1H). As we would have expected based on our hypothesis that the glucose concentration in our standard diet was too low to cause an increase in IPC activity after starvation, we find that feeding standard diet plus glucose increases IPC activity to the same level as glucose only, and that adding arabinose to the standard diet does not lead to increased IPC activity after starvation (Figure 1H).

The incretin-like effect is exciting and it will be interesting in the future to find out what might be the signal mediating this effect. It is interesting that IPCs in explants seem to be responsive to glucose. I think it would help if the authors could briefly discuss possible sources for the different findings between these in fact very different preparations. Could the the absence of the inhibitory DH44 feedback in the *ex-vivo* recordings for example play a role?

We thank the referee for this interesting point and expanded our discussion accordingly. We included that, in particular in brain explants without a VNC, the inhibitory connection we describe might be absent, as the referee suggested: ‘Previous ex vivo studies suggested that IPCs, like pancreatic beta cells, sense glucose cell-autonomously(23,24). Consistent with this, we observed an increase in IPC activity after the ingestion of glucose (Figure 2B). However, IPC activity did not increase during the perfusion of glucose directly over the brain. Importantly, the fly preparations were kept alive for several hours allowing the glucose-rich saline to enter circulation and reach all body parts. Several factors may explain the difference between ex vivo and in vivo preparations. First, in ex vivo studies, certain regulatory feedback mechanisms present in vivo could be absent. For example, the strong inhibitory input IPCs receive from DH44Ns we found would likely be absent in brain explants without a VNC. A lack of inhibitory feedback might allow for more direct glucose sensing by IPCs ex vivo, whereas in vivo, the IPC response could be suppressed by more complex systemic feedback. Second, we attempted to use the intracellular saline formulation employed in a previous ex vivo study44. However, we observed that IPCs depolarized quickly using this saline, leading to unstable recordings that did not meet our quality standards for in vivo experiments. Another possible explanation for the lack of an effect of glucose might have been that the dominant circulating sugar in flies is trehalose(70,71) which is derived from glucose. When we extended our experiments, we found that trehalose perfusion did not affect IPC activity either, strengthening the idea that IPCs do not directly sense changes in hemolymph sugar levels. Therefore, our findings suggest that, similar to mammals, IPC activity and hence, insulin release, is not simply modulated by hemolymph sugar concentration in Drosophila.’

The incretin-like effect the authors observed seems to start only after 5h which seems longer than in mammals where, as far as I know, insulin peaks around 1h. Do the authors have ideas on how this timescale relates to ingestion and glucose dynamics in flies?

We have now included the following section in the discussion to explicitly address the question of different activity dynamics in flies and mammals, but also the limitations of our electrophysiological approach in this regard: ‘We observed that IPC activity increased over a timescale of hours, which is longer compared to the fast insulin response in mammals, where insulin typically peaks within an hour of feeding(97). In flies, insulin levels rise within minutes of refeeding, followed by a drop after 30 min(20). Our experimental techniques limit our ability to capture these fast initial dynamics, since the preparation for intracellular recordings requires tens of minutes, so that we typically recorded IPC activity at least 20 min after the last food ingestion. Notably, studies in fasted mammals have shown that insulin peaks within minutes of refeeding, followed by a rapid decline, with levels stabilizing as feeding continues(98,99). We speculate a similar dynamic could be present in flies, but with our approach, we capture the steady-state reached tens of minutes after food ingestion rather than a potential initial peak.’

The authors mention "a decrease in the FV of IPC-activated starved flies even before the first optogenetic stimulation (Figure 2I),". Could this be addressed by running an experiment in darkness, only using the IR illumination of their behavioral assay?

We thank the referee for pointing out this unexpected result. We discuss this in more detail in the new version of our manuscript and expand on the reasons for not performing these optogenetic activation experiments in the dark: First, the red LED required to activate CsChrimson triggers strong startle responses in dark-adapted flies, which mask other behavioral effects, in particular subtle ones such as those observed for IPCs. The startle response is much reduced when performing experiments under low background light conditions. Second, flies, at least in our hands, do not exhibit robust foraging behavior or starvation-induced hyperactivity in the dark, which is critical for our behavioral experiments. However, we also explain in our discussion that we believe the effect of background illumination is relatively small, since flies expressing CsChrimson in OANs or DH44Ns show comparable activity levels to controls. Hence, a part of this effect is likely attributable to leak currents induced by CsChrimson expression. We would like to point out though that we are careful in our description of the IPC effect on behavior, and focus on the fact that it is considerably smaller than the effects of other modulatory neurons (DH44Ns and OANs).

The authors show an inhibitory effect of DH44 neuron activation on IPC activity. They further demonstrate that DH44PI neurons are not the ones driving this and thus conclude that "...IPCs are inhibited by DH44Ns outside the PI.". As the authors mentioned the broad expression of the DH44-Gal4 line, can they be sure that the cells labeled outside the PI are actually DH44+? If so they should state this more clearly, if not they should adapt the discussion accordingly.

We have substantially added to our discussion of this point, according to the referee’s great suggestion. In short, the broad line includes neurons that are DH44 positive and neurons that are not: ‘Notably, the DH44PINs express the DH44 peptide, as confirmed by anti-DH44 stainings(100). This also applies to a large fraction of neurons labelled in the broad DH44 driver line(100). However, a subset of neurons labelled in the broad line did not exhibit DH44 immunoreactivity(100), and might therefore not actually express the DH44 peptide. Hence, the inhibition of IPCs could be driven by neurons in the DH44 driver line that do not express DH44.’

Reviewer #3 (Public Review):

Although insulin release is essential in the control of metabolism, adjusted to nutritional state, and plays major roles in normal brain function as well as in aging and disease, our knowledge about the activity of insulin-producing (and releasing) cells (IPCs) in vivo is limited.

In this technically demanding study, IPC activity is studied in the Drosophila model system by fine in vivo patch clamp recordings with parallel behavioral analyses and optogenetic manipulation.

The data indicate that IPC activity is increased with a slow time course after feeding a high-glucose diet. By contrast, IPC activity is not directly affected by increasing blood glucose levels. This is reminiscent of the incretin effect known from vertebrates and points to a conserved mechanism in insulin production and release upon sugar feeding.

Moreover, the data confirm earlier studies that nutritional state strongly affects locomotion. Surprisingly, IPC activity makes only a negligible contribution to this. Instead, other modulatory neurons that are directly sensitive to blood glucose levels strongly affect modulation. Together, these data indicate a network of multiple parallel and interacting neuronal layers to orchestrate the physiological, metabolic, and behavioral responses to nutritional state. Together with the data from a previous study, this work sets the stage to dissect the architecture and function of this network.

Strengths:

State-of-the-art current clamp in situ patch clamp recordings in behaving animals are a demanding but powerful method to provide novel insight into the interplay of nutritional state, IPC activity, and locomotion. The patch clamp recordings and the parallel behavioral analyses are of high quality, as are the optogenetic manipulations. The data showing that starvation silences IPC activity in young flies (younger than 1 week) are compelling. The evidence for the claim that locomotor activity is not increased upon IPC activity but upon the activity of other blood glucose-sensitive modulatory neurons (Dh44) is strong. The study provides a great system to experimentally dissect the interplay of insulin production and release with metabolism, physiology, and behavior.

Weaknesses:

Neither the mechanisms underlying the incretin effect, nor the network to orchestrate physiological, metabolic, and behavioral responses to nutritional state have been fully uncovered. Without additional controls, some of the conclusions would require significant downtoning. Controls are required to exclude the possibility that IPCs sense other blood sugars than glucose. The claim that IPC activity is controlled by the nutritional state would require that starvation-induced IPC silencing in young animals can be recovered by feeding a normal diet. At current firing in starvation, silenced IPCs can only be induced by feeding a high-glucose diet that lacks other important ingredients and reduces vitality. Therefore, feasible controls are needed to exclude that diet-induced increases in IPC firing rate are caused by stress rather than nutritional changes in normal ranges. The finding that refeeding starved flies with a standard diet had no effect on IPC activity but a strong effect on the locomotor activity of starved flies contradicts the statement that locomotor activity is affected by the same dietary manipulations that affect IPC activity. The compelling finding that starvation induces IPC firing would benefit from determining the time course of the effect. The finding that IPCs are not active in fed animals older than 1 week is surprising and should be further validated.

We thank the referee for the thoughtful and constructive criticism of our experiments and conclusions. Below, we lay out how we tackled the individual points raised by the referee.

(1) ‘Controls are required to exclude the possibility that IPCs sense other blood sugars than glucose.’

To address this point, we conducted experiments in which we perfused trehalose (Figure 3B), the main circulating hemolymph sugar in Drosophila and other insects. Our results clearly show that trehalose does not affect IPC activity upon perfusion, confirming our statements that IPCs do not sense key blood sugars directly.

(2) ‘Feasible controls are needed to exclude that diet-induced increases in IPC firing rate are caused by stress rather than nutritional changes in normal ranges’.

We agree with the referee that this point was not completely fleshed out in our first submission. We have now performed additional experiments in which we added glucose (and fructose) to our standard diet (Figure 1H). Flies feeding on this diet received all necessary nutrients but still experienced high concentrations of sugars. The effects of high glucose in a standard diet background were indistinguishable from those of high glucose in agarose, confirming that the IPCs respond to sugar rather than stress. Another important observation in this context is that IPCs in flies kept on a high protein diet exhibited much lower spike rates than flies exhibiting the high glucose diet, even though they had a much shorter lifespan and therefore, presumably, experienced much higher stress levels (Figure 1H, Figure S1). These observations underline that stress is certainly not the primary factor here.

(3) ‘The finding that refeeding starved flies with a standard diet had no effect on IPC activity but a strong effect on the locomotor activity of starved flies contradicts the statement that locomotor activity is affected by the same dietary manipulations that affect IPC activity.’

We have revised the respective section of the results and discussion accordingly and are more careful and clearer in our interpretation of this behavioral dataset: ‘These results show that the locomotor activity was affected by the same dietary manipulations that had strong effects on IPC activity. However, IPC activity changes alone cannot explain the modulation of starvation-induced hyperactivity. On the one hand, high-glucose diets which drove the highest activity in IPCs were not sufficient to reduce locomotor activity back to baseline levels. On the other hand, refeeding flies with SD did not revert the effects of starvation on IPC activity (Figure 1H), but it was sufficient to reduce the locomotor activity below baseline levels (Figure 2B). This suggests that the modulation of starvation-induced hyperactivity is achieved by multiple modulatory systems acting in parallel.’

(4) ‘The compelling finding that starvation induces IPC firing would benefit from determining the time course of the effect.’

We followed the referee’s excellent suggestion and determined the time course of the starvation effect in three timesteps, similar to the experiments we did for refeeding (Figure 1G). In addition, we now also quantify the number of active IPCs (i.e., IPCs that fired at least one action potential during our five-minute analysis window), which further illustrates the dynamics of the starvation and refeeding effects. We find that the starvation effect is graded, and that IPC activity decreases with increasing starvation duration.

(5) ‘The finding that IPCs are not active in fed animals older than 1 week is surprising and should be further validated.’

To address the referee’s comment, we have added 14 new IPC recordings from flies in the 6–26-day range, such that we now have recordings from 9-14 IPCs for each age range (Figure S2B). They confirmed our previous analysis and strengthened the finding that IPC activity dramatically decreases after 8 days (on our standard diet). The total number of IPCs in this supplementary dataset was thus increased from 34 to 48.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) Do IPCs respond to glucose specifically after ingestion or generally to any other nutritive sugars? To tackle this question the IPC responses in starved flies can be recorded after refeeding flies with other nutritive sugars (fructose, sucrose).

To address this important question, we have performed additional experiments in which we refed starved flies with fructose, as a nutritive sugar, and arabinose, as a non-nutritive sugar. As expected, IPCs responded to fructose but not arabinose and hence nutritive sugars in general. We describe and discuss these key results in the new version of our manuscript.

(2) In Figure 2, the x and y axes are not annotated on all subfigures, which might help improve clarity.

We have annotated the subfigures as requested.

(3) In the discussion on page 9 ("...we observed an increase in IPC activity after the ingestion of glucose (Figure 2B)."), the authors refer to Figure 2B instead of 3C.

We have fixed this oversight.

Reviewer #2 (Recommendations For The Authors):

Introduction

I think it could be helpful for the reader if you would briefly state the number of IPCs and whether you are targeting all of them with Dilp2-Gal4.

We included the numbers according to the suggestion. 14 IPCs are labeled in the driver line, and this is the number of IPCs commonly assumed to be present in the PI.

Figures

In some Figures (for example 1D & E) the authors state the number of IPCs recorded (N) but not the number of animals used (n). This should be stated as the data from within an animal are dependent and might give insights about IPC heterogeneity.

We have compiled tables for the supplementary material (Tables S5 & S6) in which we state the number of IPCs and DH44PINs recorded and the number of different flies for each figure panel. We have recorded an average of 1.4 IPCs per fly (217 IPCs from 160 flies). We therefore expect the bias introduced by individual flies to be rather small. However, in our parallel study, we specifically investigate the heterogeneity of IPCs by maximizing the number of IPCs recorded per fly (Held M, et al. ‘Aminergic and peptidergic modulation of Insulin-Producing Cells in Drosophila’. eLife. 2024;13. doi:10.7554/ELIFE.99548.1). In the case of DH44PINs, we recorded 24 neurons in 21 flies – 1.1 neurons per fly.

- Figure 3D: There is some white visible among the cell bodies in the overlay. I assume this comes from projecting across layers rather than indicating DH44 - IPC overlap? It would help to explicitly state that.

We have added a statement to the results section, in which we explain that most of the white is due to overlap in the z-projection rather than overlap in the driver lines. However, there are few cases (typically one to two cells per brain), in which neurons labeled by the DH44 line also stain positive for Dilp2, indicating they express both neuropeptides. We have added this information to the manuscript:

Results: ‘DH44PINs are anatomically similar to IPCs, and their cell bodies are located directly adjacent to those of IPCs in the PI, making them an ideal positive control for our experiments (Figure 3D). A small subset of DH44PINs also expresses Dilp2(75), and our immunostainings confirmed colocalization of Dilp2 and DH44 in a single neuron (Figure 3D, white arrow).’

In figure caption: ‘UAS-myr-GFP was expressed under a DH44-GAL4 driver to label DH44 neurons. GFP was enhanced with anti-GFP (green), brain neuropils were stained with anti-nc82 (cyan), and IPCs were labelled using a Dilp2 antibody (magenta). White arrow indicates Dilp2 and DH44-GAL4 positive neuron. The other white regions in the image result from an overlap in z-projections between the two channels, rather than from antibody colocalization.’

- Figure 4I: One might get the impression that the fast onset peak of activity precedes the stimulation onset, using a thinner line width might help avoid that.

This effect is due to a combination of using relatively heavy lines for clear visibility of the data and a gentle smoothing step (a 2s median filter, which corresponds to less than 1% of the 300s stimulation window) in our analysis of the behavioral data. However, inspection of the raw data clearly shows increases in velocity after the onset of the optogenetic activation. We clarified this in the figure caption: ‘Average FV across all DH44N activation trials based on two independent replications of the experiment in I. Note that the peak in average FV lies within the first frame of the stimulation window.’

- S3 panel letters do not match references in the text.

We fixed this oversight.

Formatting

- Page 10: The paragraphs on the bottom of the page got switched around.

This has been fixed.

- Page 14: The first paragraph after the header "Free-walking assay" seems to be coming from elsewhere.

We apologize for this slightly embarrassing mistake. We used our related bioRxiv preprint (Held et al.) as a template for formatting this paper, and accidentally left this part of the methods section in the manuscript. We have fixed this error in our resubmission.

Reviewer #3 (Recommendations For The Authors):

Major suggestions:

(1) The data show convincingly that IPC activity is decreased by starvation during the first week of adult life (Figures 1C and D). However, the conclusion that IPC activity is controlled by the nutritional state requires additional care. First, refeeding starved adult animals with a normal diet does not bring back normal IPC firing rates (Figure 1H). Therefore, IPC activity does not strictly follow changes in nutritional state, but IPCs are silenced by starvation. Second, from the second week of adult life on, IPCs are silent anyway, and thus unlikely responsive to changes in the nutritional state anymore (which might be different on a different standard diet?) The only effect of feeding on IPC activity is observed upon feeding starved, young animals with high glucose for 12-24 hrs (Figure 1G). However, it is not clear whether increased IPC firing is caused by the effects of high glucose on the nutritional state in a normal range, or because of diet-induced stress (the diet also severely shortens lifespan, Figure 1S). Does high glucose also increase IPC firing rate in young, fed animals? These would have strongly increased glucose concentrations but not suffer the stress of not getting any other nutrients. Such experiments would be required to make the statement that glucose feeding increases IPC firing rate.

We have performed several experiments to address this criticism. First, we performed a time course analysis of the starvation effect. We show that the IPC activity reduction is graded, and that IPC activity declines already after two hours of starvation, a timepoint at which stress levels should still be relatively small (Figure 1G). Second, we refed flies with high glucose concentrations added to the standard diet (Figure 1H). This minimized any potential stress responses due to a lack in nutrients. Third, we now show that IPCs specifically respond to nutritive (glucose and fructose), but not to non-nutritive sugars (arabinose, Figure 1H). We believe that these data sets, in addition to the graded refeeding effect, make a strong case for the nutritional state dependent modulation of IPCs.

(2) The testing of locomotor activity is well done, nicely recapitulates starvation-induced increases in locomotion, and adds interesting novel findings on refeeding with high glucose versus high protein diet. However, the statement that locomotor activity was affected by the same dietary manipulations that had strong effects on IPC activity does not reflect the data presented. Refeeding starved flies with a standard diet had no effect on IPC activity (Figure 1H) but a strong effect on locomotor activity of starved flies (a strong reduction, even stronger than high glucose diet, Figure 2B).

We have revised the respective section of the results and discussion accordingly and are more careful and clearer in our interpretation of this behavioral dataset: ‘These results show that the locomotor activity was affected by the same dietary manipulations that had strong effects on IPC activity. However, IPC activity changes alone cannot explain the modulation of starvationinduced hyperactivity. On the one hand, high-glucose diets which drove the highest activity in IPCs were not sufficient to reduce locomotor activity back to baseline levels. On the other hand, refeeding flies with SD did not revert the effects of starvation on IPC activity (Figure 1H), but it was sufficient to reduce the locomotor activity below baseline levels (Figure 2B). This suggests that the modulation of starvation-induced hyperactivity is achieved by multiple modulatory systems acting in parallel.’

Related to points 1 and 2, a key statement that the results establish that IPC activity is controlled by the nutritional state requires care. What the data convincingly show is that IPC activity is near zero upon starvation.

As described above, we have added several extensive data sets (fructose feeding, arabinose feeding, trehalose perfusion, starvation time course) to show that we indeed observe a nutritional state dependent modulation of IPCs and describe these new results in the results and discussion.

(3) The time course of nutritional state-dependent changes of IPC activity is claimed to be slow, several hours to days. Unless I have missed a figure, the underlying data are not presented (only for high glucose diet). It would be great if this could also be shown for a standard diet with higher glucose concentrations than the one used so that it rescues starvation-induced IPC silencing without shortening lifespan (if this is feasible?). The data showing starvation-induced IPC silencing are convincing, but, unless I have missed it, the time course has not been determined. It would be very nice to actually show this. Have different starvation times been tested in relation to IPC firing rate, and if yes, with what time resolution? Does IPC activity change already after 0.5 or 1 or a few hours of starvation? If starvation can silence IPCs faster than assumed, the nearzero IPC activity in animals older than a week could very well be caused by longer time intervals between meals.

We have performed experiments to address both important points raised by the referee here. 1) We have added high glucose concentrations to our standard diet, and show that it has the same effect – a significant increase in IPC activity – as the high glucose diet (Figure 1H). 2) We have analyzed the time course of IPC activity reduction in response to starvation (Figure 1G). Indeed, we find that a few hours of starvation start reducing IPC activity. We discuss the possibility that reduced IPC activity in older flies could be due to reduced food intake: ‘One of our experiments demonstrated that IPC activity was heavily diminished in flies older than 10 days (Figure S2B). A possible explanation could be that flies feed less as they age. However, this only holds true for flies older than 14 days86. Therefore, reduced IPC activity in 10-11 day old flies is unlikely to result from reduced food intake and likely involves inhibition of insulin signaling.’

(4) The data on the proposed incretin effect are of high importance in potentially highlighting a highly conserved link between glucose ingestion and insulin release. An important control would be to test different sugars, such as trehalose, an important blood sugar of flies. If glucose is converted into trehalose and this is what IPCs sense, then perfusion of glucose has no effect. The fact fantastic experiments show that the DH44 neurons are sensitive to glucose perfusion does rule out that IPCs sense a different sugar. This would be very different from the incretin effect that requires additional hormones. In addition, as mentioned above, controls are required to show that high glucose affects IPCs as a nutrient and not as a stressor (see point 1), for example refeeding with a standard diet that contains a higher glucose concentration but does not reduce lifespan. Another great control to solidify the exciting claim on the incretin effect would be to knock out candidate Drosophila incretin hormones and test whether a high glucose diet stops increasing the IPC firing rate (although simpler controls might also do the job).

We have performed the two key experiments suggested by the referee. 1) We perfused trehalose as the primary blood sugar of flies and showed that IPCs do not respond to trehalose perfusion (Figure 3B & C). This further strengthens the finding that IPC activity in flies shows an incretin-like effect. 2) We have added high concentrations of glucose to our standard diet to provide flies with a full diet that contains high glucose concentrations. IPC activity in these flies was indistinguishable from the activity in flies which consumed pure glucose diets. In contrast, IPC activity in flies kept on a high protein diet, which dramatically reduced lifespan, was very low. These results clearly show that higher IPC activity is not due to increased stress levels, but a function of nutritive sugar ingestion. We further validated this hypothesis by refeeding flies with fructose as a nutritive sugar, which increased IPC activity, and arabinose as a non-nutritive sugar, which did not affect IPC activity (Figure 1H).

Another point that might be relevant to this discussion is that IPC activity is almost entirely shut down during flight in Drosophila (which we showed in Liessem et al. 2023, Current Biology 33 (3), 449-463. e5). Several ‘stress hormones’ are released during flight, including octopamine. The fact that IPC activity is low in flying flies, starved flies, and flies kept on a pure protein diet (which all experience high stress levels), to us, very clearly suggests that stress is not the predominant factor here. We would also like to point out that, while the lifespan was reduced in flies kept on pure glucose diets, survival rates were at 100% until day 14, and we carried out our experiments on day 2 after starvation. Hence, these flies might not (yet) experience particularly high stress levels.

(5) The discussion relates the absence of IPC firing in animals older than 1 week to aging. However, given that the flies fed on a normal diet show the typical lifespan for Drosophila, a 10-dayold fly is still in its youth. Maybe flies at 10 days eat simply less and thus IPC spiking goes down as in starved flies, especially because the standard diet used contains low glucose. Do IPCs also become silent after a week if the animals are fed with a standard diet that contains a higher glucose concentration? Without additional controls, this part of the discussion is pretty speculative and should be revised.

We agree with the reviewer, that it is not clear whether reduced IPC activity is a direct result of physiological changes that occur with aging, or an indirect effect of reduced food intake, which occur during aging. In both cases, in our view, it would be an age-related effect. Since this is a minor point of our manuscript, we decided not to perform additional experiments, other than significantly increasing the sample size for the aging data set already presented to shore up our findings (Figure S2B). We have, however, revisited the discussion of this point according to the referee’s suggestion: ‘One of our experiments demonstrated that IPC activity was heavily diminished in flies older than 10 days (Figure S2B). A possible explanation could be that flies feed less as they age. However, this only holds true for flies older than 14 days(85). Therefore, reduced IPC activity in 10-11 day old flies is unlikely to result from reduced food intake and likely involves inhibition of insulin signaling.’

Other suggestions:

(6) For the mixed effects of octopamine and tyramine on larval locomotion that are referred to, it might be interesting to also look at Schützler et al 2019, PNAS because it shows that starvation activates TBH so that the octopamine to tyramine ratio is increased.

We refer to Schützler et al. in the following paragraph of our discussion: ‘This intermittent locomotor arrest has been previously described in adult flies and is thought to be mediated by ventral unpaired median OANs, which have been suggested to suppress long-distance foraging behavior(69). Since these are not the only neurons we activate in the TDC2 line, we speculate that the stopping phenotype could also result from concerted effects of octopamine and tyramine modulating muscle contractions(65-67) and motor neuron excitability(68), as previously described in Drosophila larvae, or from OANs interfering with pattern generating networks in the ventral nerve cord (VNC) during longer activation(69).’

(7) The reference list requires care. For example, reference 43 is identical to 67, reference 66 gives no information on incretin-like hormones in Drosophila as stated in the text

We carefully double-checked our reference list and corrected the mistakes mentioned.

Version 2

Abstract

Insulin plays a key role in metabolic homeostasis across vertebrate and invertebrate species. Drosophila Insulin-Producing Cells (IPCs) are functional analogues to mammalian pancreatic beta cells and release insulin directly into circulation. IPC activity is modulated by nutrient availability, circadian time, and the behavioral state. To investigate the in vivo dynamics of IPC activity in the context of metabolic homeostasis, we quantified effects of nutritional and internal state changes on IPCs using electrophysiological recordings. We found that the nutritional state strongly modulates IPC activity. IPCs became less active with increasing periods of starvation. Refeeding starved flies with glucose or fructose, two nutritive sugars, significantly increased IPC activity, whereas non-nutritive sugar and protein had no effect. In contrast to feeding, glucose perfusion did not affect IPC activity. This was reminiscent of the mammalian incretin effect, in which ingestion of glucose drives higher insulin release than intravenous glucose application. Contrary to IPCs, Diuretic hormone 44-expressing neurons in the pars intercerebralis (DH44PINs), which are anatomically similar to IPCs, responded to glucose perfusion. Functional connectivity experiments demonstrated that glucose-sensing DH44PINs do not affect IPC activity, while other DH44Ns inhibit IPCs. This suggests that populations of autonomously and systemically sugar-sensing neurons work in parallel to maintain metabolic homeostasis. Ultimately, metabolic state changes affect behavior. For example, hungry flies increase their locomotor activity in search of food. In support of this idea, activating IPCs had a small, satiety-like effect on starved flies, resulting in reduced walking activity, whereas activating DH44Ns strongly increased walking activity. Taken together, we show that IPCs and DH44Ns are an integral part of a sophisticated modulatory network that orchestrates glucose homeostasis and adaptive behavior in response to shifts in the metabolic state.

Introduction

Insulin is a key neuropeptide that controls metabolic homeostasis across the animal kingdom1–4. Beyond metabolic homeostasis, insulin signaling is implicated in physiological processes underlying reproduction, aging, and stress resistance5–7. When nutrients are abundant, insulin promotes the uptake and storage of energy. Conversely, during periods of starvation, insulin release is downregulated8,9, which promotes the conservation of energy resources and enables energy mobilization via counter-regulatory pathways10–12. Maintaining this energy balance is vital for survival. Accordingly, dysregulation of this finely tuned system is a hallmark of metabolic disorders such as type 2 diabetes13–15.

Drosophila melanogaster has emerged as a model system to study insulin signaling because of the availability of a powerful genetic toolkit16–21. In Drosophila, 14 Insulin-Producing Cells (IPCs), which are analogous to human pancreatic beta cells, reside in the pars intercerebralis (PI) of the brain22. These IPCs secrete Drosophila insulin-like peptides (DILPs) and are sensitive to changes in the metabolic state23–26, the behavioral state27, the circadian time25,28, aminergic and peptidergic modulatory inputs29,30, and other peripheral signals31. Consequently, they play a key role in orchestrating the metabolic demands of behaving animals. For instance, insulin acts as a global cue for modulating odor-driven food-search in accordance with the fly’s nutritional state32. Unlike pancreatic beta cells, IPCs are localized in the brain and engage in direct synaptic signaling30,33 in addition to endocrine signaling. Thus, IPCs are part of a complex system that integrates both neuronal and peripheral signals. As a key hub integrating metabolic state information, IPCs are thought to be part of the neuronal circuitry that regulates food intake32,34–39, in part by regulating foraging behavior. Previous studies found that octopamine, the insect analogue to norepinephrine, mediates starvation-induced hyperactivity in flies38. A subset of octopaminergic neurons expressing Drosophila insulin receptor (dInR) are necessary for the regulation of starvation-induced hyperactivity38,40 implicating that IPCs play a role in modulating foraging behavior.

While much is known about carbohydrate metabolism and insulin signaling in Drosophila5,41–43, the in vivo dynamics of IPCs remain largely elusive. Ex vivo studies indicate that IPCs themselves are sensitive to glucose, as evidenced by increased firing rates upon glucose perfusion23,44. However, the influence of feeding and nutritional state changes on IPC activity in vivo, as well as the nutritional state-dependent response of IPCs to glucose, remain unclear, underscoring the importance of in vivo studies where all nutrient-sensing and sensorimotor circuits are intact. For instance, how does IPC activity differ when glucose is ingested as part of a meal, compared to when glucose is administered directly to the brain or into circulation? In mammalian systems, insulin release is significantly amplified upon glucose ingestion vs. intravenous administration45,46, a phenomenon known as the ‘incretin effect’47,48. This underlines the complexity of insulin regulation, emphasizing that insulin secretion is not solely dictated by sugar levels but influenced by multiple factors. Here, we investigated the nutritional state-dependent modulation of IPCs in Drosophila using an in vivo electrophysiological approach.

We recently demonstrated that IPC activity is strongly modulated by locomotion on fast timescales27. Whether the converse is true, and IPC activity can directly affect behavior on short timescales, is not clear. To address this, we optogenetically manipulated IPCs and other modulatory neurons while recording the locomotor activity of freely walking flies.

In combination, our approaches show that IPC activity is not just nutritional state-dependent, but that IPCs also exhibit an incretin-like effect. We further establish that the nutritional state strongly modulates locomotor activity, and that IPCs play a small but significant role in this. In addition, we confirm that another set of modulatory neurons adjacent to the IPCs, DH44PINs, are sensitive to glucose perfusion. While DH44PINs do not affect IPC activity on short timescales, DH44 neurons outside the PI (DH44Ns) form strong inhibitory connections to IPCs. These DH44Ns also have strong effects on locomotor activity, which are antagonistic to those of IPCs. Our findings imply distinct mechanisms by which IPCs and DH44PINs sense changes in nutritional state, underlining the significance of multiple pathways in orchestrating Drosophila’s response to nutritional state changes and ensuring metabolic homeostasis.

Results

IPC activity is modulated by starvation and glucose feeding

To characterize the electrophysiological activity of IPCs in different nutritional states, we performed in vivo patch-clamp recordings from IPCs in fed and starved Drosophila (Figure 1A-C). First, flies were fed ad libitum, and the IPC baseline activity was determined by recording their spontaneous activity in glucose-free extracellular saline in a five min window. The baseline firing rate varied between individual IPCs and ranged from 0 to 1.4 Hz (Figure 1D). To assess the effect of starvation on IPC activity, we wetstarved flies for 24 h. IPCs were significantly less active in starved flies than in fed flies with a median firing rate of 0 Hz (Figure 1D). Hence, the IPC population was basically quiescent after starvation. We also analyzed the resting membrane potential (Vm) of IPCs in fed and starved flies. On average, the Vm of IPCs was 8 mV more depolarized (median: –51 mV) in fed flies compared to starved flies (median: –59 mV, Figure 1E). This indicates a higher excitability of IPCs after feeding, which likely explains their higher firing rates.

IPC activity is modulated by the nutritional state and increases in response to feeding on nutritive sugars.

A) Schematic of the setup for in vivo IPC whole cell patch-clamp recordings. B) IPCs in the Drosophila brain. UAS-myr-GFP was expressed under a Dilp2-GAL4 driver to label all 14 IPCs. The GFP signal was enhanced with anti-GFP (green), brain neuropils were stained with anti-nc82 (magenta). C) Representative examples of the membrane potential of two IPCs recorded in fed (magenta) and starved (cyan) flies. D) Average baseline spike rate and E) membrane potential of IPCs in fed (magenta) and starved (cyan) flies. F) Schematic of the experimental starvation and refeeding protocol. HG: High glucose, HG+SD: High glucose with standard diet, HF: High fructose, HA+SD: High arabinose with standard diet, SD: Standard diet, HP: High protein. G) Comparison of IPC spike rates in fed flies (magenta), increasingly starved flies (cyan), and flies refed on HG for different durations (green). Right axis shows fraction of active IPCs (number of IPCs with spike rate > 0 Hz, gray lines and circles). H) Comparison of IPC spike rate in flies refed on different diets. Reference lines represent median IPC activity in flies refed with HG for 18-24 h (green), and in flies starved for 24 h (cyan). Each circle represents an individual IPC, N = number of IPCs (see table S5 for number of flies), error bars indicate median (circle) and interquartile range (IQR, bars). For D), E) and G), p-values are reported from Wilcoxon rank-sum test. For H), a Kruskal-Wallis test followed by post hoc Wilcoxon rank-sum tests were used for pairwise comparisons. p-values are reported after Holm-Bonferroni correction.

To explore the temporal dynamics of starvation-induced decreases in IPC activity, we recorded IPCs in flies subjected to increasing durations of starvation. The IPC population activity decreased steadily with increasing starvation durations and reached a median of 0 Hz after 24 h (Figure 1G). The fraction of active IPCs, that is, IPCs that fired at least one action potential during our five min recording window, also progressively decreased with increasing periods of starvation (Figure 1G, gray). The reduction in IPC activity was already noticeable after 0.5-2 h of starvation, suggesting that starvation tightly regulates IPC activity and hence insulin release on shorter and longer timescales.

Having established that starvation reduces IPC activity, we next explored if IPC activity could be restored by feeding. Therefore, we first starved flies for 24 h and then refed them with a high glucose (HG) diet (400 mM glucose in 2% agar) for durations ranging from 3 to 24 h (see Figure 1F and 1G). IPC activity in flies that were refed for 3-5 h resembled that of starved flies and remained around 0 Hz (Figure 1G). 6 h after the start of refeeding, IPC activity and the fraction of active IPCs started to increase and were similar to that of fed flies (Figure 1G, p = 0.68), indicating that this duration was sufficient to allow IPC activity to fully recover from starvation. Interestingly, IPC activity kept increasing for longer glucose feeding durations, and was significantly increased in flies that were refed with glucose for 18-24 h compared to flies kept on a standard diet and fed ad libitum (Figure 1G, p = 0.02). Together, these results clearly show that IPC activity decreases with increasing duration of starvation up to 24 h, and slowly but steadily recovers and eventually overshoots the baseline after refeeding with glucose.

To ensure that the results obtained on an HG diet were not due to a lasting interference with the fly’s metabolic system, we conducted a survival assay. Analogous to our electrophysiology experiments, we starved flies for 24 h and refed them on HG. Subsequently, we monitored the number of survivors every day while flipping flies onto fresh food every second day. As expected, flies refed with our standard diet (SD) survived the longest (100% survival until day 43, Figure S1). 100% survival was observed for flies fed with HG until day 12 and at least half of the flies survived until day 16 (Figure S1). Since our IPC recordings were acquired in flies refed for a maximum of 24 h, the effects we observed on IPC activity were not due to lethality in flies but rather due to differences in nutrient availability

IPCs are sensitive to ingestion of nutritive sugars

The survival assay indicated that the HG diet was stressful for flies, as it did not contain all necessary nutrients. To ensure that the increased IPC activity was not a stress response but due to glucose ingestion, we added 400 mM glucose to our standard diet (HG+SD) and reassessed the effect of glucose ingestion in a full diet background. Refeeding flies with HG+SD caused a significant increase in IPC activity (Figure 1H), which was indistinguishable from that of HG only (Figure 1H), confirming that IPCs respond to glucose feeding.

Next, we tested whether IPC responses are glucose specific – in principle, IPCs could also be sensitive to the ingestion of other nutritive sugars. To this end, we kept flies on a high fructose diet (HF, 400 mM Fructose in 2% agar), which is a natural source of nutritive sugar for flies, found in fruit. IPCs in flies kept on the HF diet exhibited a similar activity level as in flies kept on HG (Figure 1H), suggesting IPCs generally respond to the ingestion of nutritive sugars. To confirm this hypothesis, we investigated if IPCs respond to a non-nutritive but sweet-tasting sugar. Hungry flies are motivated to feed on non-nutritive sugars, and the sweet taste or sugar concentrations may affect insulin signaling. To test this, we refed flies with 400 mM D-arabinose. Since arabinose is non-nutritive, we supplied it in our standard diet (HA+SD). In contrast to nutritive sugars, refeeding flies with D-arabinose did not increase IPC activity after starvation. Accordingly, IPC activity was significantly lower than in flies fed with HG or HF, and indistinguishable from starved flies (Figure 1H). These results clearly show that IPC activity is specifically increased in response to ingestion of nutritive sugars.

It was unexpected that IPC activity after refeeding with the HA+SD diet was comparable to starved flies, since we expected it to recover to levels observed in flies fed ad libitum (Figure 1G, magenta). However, refeeding flies on SD only, without additional sugars, did not increase IPC activity compared to the 24 h starved state, even though we refed flies for 18–24 h, where we saw the maximal effect of glucose refeeding (Figure 1H). This was likely because sugar concentrations in our SD were low (see Discussion). To further confirm whether IPC activity increases after refeeding were sugar-specific or could also be recovered by other specific nutrients, like protein, we kept 24 h starved flies on a high protein diet (HP) and compared the IPC activity after refeeding for 18-24 h. Similar to refeeding on SD, refeeding on an HP diet caused no significant increase in the spike rate of IPCs (Figure 1H). This was intriguing, because both protein-rich diets and standard diets have been shown to induce insulin secretion under fed conditions26,49,50. In comparison to HG and SD, flies survived poorly on the HP diet: 100% survival was observed only up to day one, and only 43% of the flies survived longer than day 12 (Figure S1). We can therefore not rule out the possibility that flies on HP exhibited a significantly impaired metabolism. Overall, our refeeding experiments demonstrate that IPCs specifically respond to feeding on nutritive sugars and exhibit relatively slow temporal dynamics on the scale of hours in this context.

Since IPCs are central to metabolic regulation and development, factors other than the nutritional state may impact their activity. To address two key factors, we analyzed whether IPC activity was affected by the mating state or the age of the fly. Although median IPC activity was the highest in males, there were no significant differences between virgin females, mated females and male flies Figure S2A). Hence, IPC activity is not significantly affected by sex or female mating state. However, when recording the IPC activity in females of different age groups, we observed that IPCs remained almost quiescent in fully fed females older than ten days. Here, baseline spike rates ranged from 0–0.4 Hz and were significantly lower than in younger females (Figure S2B). This indicates that IPC activity is reduced in older flies. Therefore, aging significantly reduces IPC activity.

The nutritional state affects locomotor activity

Starvation increases baseline locomotor activity in flies and other animals. This ‘starvation-induced hyperactivity’ is a hallmark of foraging behavior32,38,40,51–54 as it enables animals to explore their environment and locate food sources. As major integrators of metabolic state information, IPCs are thought to be part of the neuronal circuitry that modulates foraging40. For example, insulin inhibits a subset of olfactory sensory neurons, such that decreased insulin release leads to an increase in sensitivity to food odors55. Our findings that IPC activity is directly modulated by starvation and feeding support the hypothesis that they could be part of the neuronal networks modulating foraging. To further investigate the relationship between the nutritional state, IPC activity, and for-aging behavior, we used our ‘Universal Fly Observatory’ (UFO) setup to quantify locomotor activity56, in particular starvationinduced hyperactivity (Figure 2A). In the UFO, flies can walk freely in a circular arena illuminated by infrared LEDs, and their walking activity can be quantified using semi-automated tracking software and analysis (based on the Caltech Fly Tracker57, see Methods). To test whether IPCs play a causal role in modulating starvation-induced hyperactivity, we first quantified the effects of nutritional manipulations on walking speed. Fully fed flies, which have elevated IPC activity, displayed a low baseline walking activity with a median forward velocity (FV) of 0.3 mm/s (Figure 2B). 24 h starved flies, in which IPC activity was significantly reduced, were more hyperactive and displayed a significantly increased median FV of 1.9 mm/s (Figure 2B). Refeeding flies on a pure glucose diet for 24 h after starvation, which led to a drastic increase in IPC activity (see Figure 1G), reduced the hyperactivity by about 50% (median FV = 0.9 mm/s) compared to starved flies (Figure 2B). Hence, the hyperactivity was significantly reduced but not abolished, despite the fact that IPC activity was maximal in these flies (Figure 1G). One interpretation of the persistent increase in activity after glucose refeeding is that flies are searching for protein sources. Flies refed on a full diet (SD) were even less active than flies fed ad libitum before undergoing starvation, and this difference was significant (Figure 2B). These results show that the locomotor activity was affected by the same dietary manipulations that had strong effects on IPC activity. However, IPC activity changes alone cannot explain the modulation of starvation-induced hyperactivity. On the one hand, high-glucose diets which drove the highest activity in IPCs were not sufficient to reduce locomotor activity back to baseline levels. On the other hand, refeeding flies with SD did not revert the effects of starvation on IPC activity (Figure 1H), but it was sufficient to reduce the locomotor activity below baseline levels (Figure 2B). This suggests that the modulation of starvation-induced hyperactivity is achieved by multiple modulatory systems acting in parallel.

Walking activity is modulated by the nutritional state, IPCs and OANs.

A) Schematic showing the UFO setup. B) Average forward velocity (FV) of flies in different feeding states. Median and IQR are shown. p-values from Wilcoxon rank-sum test. C) Average FV of flies representing one replicate during optogenetic activation of IPCs in fed flies (magenta). Empty split-GAL4 was used as control for all experiments (black). Red shading, optogenetic activation (300 s). D) Average FV across all trials from two replicates for IPC activation in fed flies: 300 s before stimulus onset, during stimulus, and after stimulus offset. N = number of flies, n = number of activation trials. Thin lines represent individual trials, thick lines represent median of all trials. E) Average FV of all flies during each stimulus trial (1-5) and poststimulus trial (300 s window immediately after activation ceased, P1-P5) for IPC activation in fed flies. F) Average FV of flies representing one replicate during optogenetic activation of IPCs in starved flies (cyan). G) Average FV across all trials from two replicates for IPC activation in starved flies (plot details as in D). H) Average FV during individual trials for IPC activation in staved flies (plot details as in E). I) Average FV of flies representing one replicate during optogenetic activation of OANs in fed flies (lilac). J) Average of FV across all trials from two replicates for OAN activation in fed flies (plot details as in D). K) Average FV during individual trials for OAN activation in fed flies (plot details as in E). L) and M) Average FV pooled across all stimulus (1-5) and post-stimulus trials (P1-P5), respectively. Median and IQR are shown. p-values are reported from Wilcoxon rank-sum test. Where no detailed p-value is stated, asterisks represent a significant difference. See also tables S1 and S2.

To develop a deeper understanding of the nutritional state-dependent modulation of behavior, we investigated the temporal dynamics of the changes in locomotor activity. Since the nutritional state-dependent modulation of IPC activity occurred on a relatively slow timescale over several hours to days, we tested whether starvation-induced hyperactivity was modulated on a similar timescale. To this end, we quantified the walking activity of flies that were either fed ad libitum or starved for increasing durations, ranging from 30 min to 48 h. As established earlier, flies that were fully fed or merely starved for 30 min had a low FV of 0.3 mm/s (Figure S3A). However, their walking activity increased with increasing starvation duration up to 20 h, when it reached a maximum at an FV of 3.6 mm/s (Figure S3A). The FV of flies starved for > 20 h decreased again, probably due to a lack of energy reserves. We observed a final peak after 36 h of starvation, which we interpret as a surge in starvation-induced hyperactivity before the metabolic state becomes critical (Figure S3A). These results demonstrate that the starvation duration has a strong effect on locomotor activity, the modulation of which occurs on a similar timescale as IPC modulation by changes in nutritional state.

IPC activation slightly reduces starvation-induced hyperactivity

Having confirmed a strong effect of starvation on both IPC activity and locomotor activity, our next goal was to investigate whether changes in IPC activity contribute to the control of starvation-induced hyperactivity. To achieve this, we optogenetically activated IPCs via CsChrimson58 in freely walking flies, which were either fed ad libitum or wet-starved for 24 h. In parallel, we used empty split-GAL459 flies crossed to UAS-CsChrimson as controls. Controls and experimental flies were simultaneously exposed to the same activation protocol in adjacent UFO setups. This approach allowed us to account for activity changes due to the red light used for optogenetic activation and other factors potentially affecting walking activity, such as small fluctuations in temperature, humidity, or differences in the circadian time. Each optogenetic activation experiment consisted of five activation cycles, in which we activated the neurons for 300 s, followed by an inter-stimulus-interval of 600 s for recovery, which we refer to as the ‘post activation’ window. Results reported here were analyzed by averaging two replicates per experiment.

IPC activation in fed flies had only minor effects on the FV (Figure 2C-E). The FV of most flies remained below 2 mm/s throughout the experiment (75th percentile, Figure 2E: see 1-5 and P1-P5) and was indistinguishable from the FV of controls during all but the first activation cycle (Figure 2C and 2E: p > 0.05). Accordingly, their FV did not differ during most activation windows (median (1-5) = 0-0.5 mm/s, Figure 2E) or post-activation windows (median (P1-P5) = 0 – 0.1 mm/s). One exception was the first activation cycle, in which the IPC-activated flies walked slightly but significantly faster than controls (Figure 2E, 1: p = 0.03). When averaging the activity of all flies across all activation windows, there was a small but statistically significant difference of 0.06 mm/s (Figure 2L, p = 0.01). In the post-activation windows, however, the FV of IPC activated and control flies were indistinguishable (Figure 2M). Overall, IPC activation had no notable effects on locomotor activity in fully fed flies.

Next, we tested the effect of IPCs on starvation-induced hyperactivity by activating IPCs in starved flies. Given that higher IPC activity levels are associated with the fed state, we expected that activating IPCs would reduce starvation-induced hyperactivity. As previously observed, starved flies were generally more active than their fed counterparts and walked with a median FV > 2 mm/s (Figure 2F-H). The FV of flies during IPC activation remained significantly lower than in control flies over the course of all five activation cycles (Figure 2F-H, p < 0.05, Table S1). This confirmed our hypothesis that IPC activation reduces starvationinduced hyperactivity. However, two observations were counterintuitive: First, the FV of IPC-activated flies and control flies converged over time, and there was no significant difference in the control and IPC-activated flies during the last two post activation windows (Figure 2F, and 2H: P4, p = 0.23 and P5, p = 0.45, see also Table S2). Hence, while activation of IPCs reduced hyperactivity, this effect did not persist after long-term activation. Second, we noted a decrease in the FV of IPC-activated starved flies even before the first optogenetic activation (Figure 2F), possibly due to CsChrimson sensitivity to visible light (see methods). Overall, these results suggest that IPCs have a small but significant effect on behavior, in that they slightly reduce starvationinduced hyperactivity upon activation.

Activation of octopaminergic neurons increases locomotor activity

Since changes in IPC activity alone were insufficient to explain starvation-induced hyperactivity, we next investigated octopamine, a neuromodulator known to induce hyperactivity in starved flies40. Octopamine is the insect analogue to norepinephrine and is released during flight, a metabolically very demanding behavior in Drosophila and other insects60–62. Hence, we expected octopaminergic neurons (OANs) to drive an increase in locomotor activity, rendering them a useful positive control. We used a tyrosine decarboxylase 2 (TDC2) driver line to acutely activate OANs via CsChrimson58 in the UFO while tracking the walking activity of flies. TDC2 is required for octopamine synthesis from tyramine63 and TDC2-GAL4 labels all neurons producing octopamine and tyramine. Experiments were carried out following the same approach as for IPCs, and we again ran empty controls in parallel to OAN activation. We observed a strong, significant increase in FV from 0.23 mm/s (median) before OAN activation to 3.68 mm/s after five activation cycles, with a strong FV increase after each activation (Figure 2I and 2K, p < 0.001, see Table S1 and S2). Notably, the final FV reached after five cycles of OAN activation was comparable to that after 24 h of starvation (compare Figure 2I to starved controls in Figure 2F). This suggests that OAN activity can explain a large proportion of the starvation-induced hyperactivity. Interestingly, we observed a dramatic drop in FV at the onset of OAN activation (Figure 2J and 2K), during which the median FV dropped by 2 mm/s and approached 0 mm/s. This reduction in FV was due to recurring stopping behavior observed during acute OAN activation (Figure S3B, video S1). This intermittent locomotor arrest has been previously described in adult flies and is thought to be mediated by ventral unpaired median OANs, which have been suggested to suppress long-distance foraging behavior64. Since these are not the only neurons we activate in the TDC2 line, we speculate that the stopping phenotype could also result from concerted effects of octopamine and tyramine modulating muscle contractions65–67 and motor neuron excitability68, as previously described in Drosophila larvae, or from OANs interfering with pattern generating networks in the ventral nerve cord (VNC) during longer activation69. Overall, long-term optogenetic activation of OANs had a strong and significant effect on walking activity and mimicked the starvation-induced hyperactivity (Figure 2F and 2H, black and gray controls). We conclude that starvation-induced hyperactivity is mildly affected by IPCs, but predominantly controlled by populations of modulatory neurons other than IPCs, including OANs.

IPCs do not sense changes in extracellular glucose levels in vivo

Our experiments established that IPC activity is modulated by the nutritional state, and that IPCs are sensitive to glucose ingestion, similar to pancreatic beta cells. Like beta cells, IPCs have been suggested to directly sense changes in extracellular glucose concentration in ex vivo studies23,44, where IPC activity increased in response to the application of saline containing a high glucose concentration to the nervous system. Consequently, IPCs are thought to sense extracellular glucose concentration increases cellautonomously. To test this hypothesis in flies with fully intact sensory and central circuits, we recorded from IPCs in vivo, while perfusing extracellular saline containing high concentrations of glucose over the brain. After recording the IPC baseline activity in starved flies under perfusion of glucose-free saline, we perfused the brain with saline containing 40 mM glucose (Figure 3A). Similar to our previous experiments, IPC spike rates were very low with a median of 0.05 Hz after starvation. Surprisingly, perfusing high concentrations of glucose over the brain did not increase IPC activity in starved flies (Figure 3A), even after 20 min of perfusion. This was similar in fed flies (Figure S4A). Towards the end of each recording session, we injected current to establish that the IPCs were still excitable, as evidenced by an increase in spike rate.

IPCs are not sensitive to glucose perfusion but DH44PINs are.

A) IPC spike rate and delta spike before and after perfusing extracellular saline with high glucose concentration, in starved flies. Baseline spike rates were recorded in glucose-free saline, followed by recordings in glucose-rich saline. Spike rates were averaged within a 5 min window. Delta spike rate was calculated by subtracting the baseline (Pre) from each trial for each IPC. Pre: 5 min recording in glucose-free saline. Glucose-rich saline was allowed to perfuse for about 8 min before analyzing IPC activity. Glu1 and Glu2: Two subsequent, 5 min long recordings in glucose-rich saline, starting eight minutes after onset of saline perfusion. Each circle represents an individual IPC from a different fly, the thick line represents the grand mean of all recordings. p-values are reported from Wilcoxon signed-rank test. B) Same experiment as in A), but with perfusion of trehalose-rich saline after recording the baseline. Note that our glucose-free saline was also devoid of trehalose (see methods). Other details same as A). C) Comparison of IPC spike rate between starved, glucose-refed, glucose-perfused and trehalose-perfused flies, highlighting the ‘incretin effect’. Median and IQR are indicated. D) Staining showing Drosophila brain with IPCs (magenta) and DH44Ns (green). UAS-myr-GFP was expressed under a DH44-GAL4 driver to label DH44 neurons. GFP was enhanced with anti-GFP (green), brain neuropils were stained with anti-nc82 (cyan), and IPCs were labelled using a Dilp2 antibody (magenta). White arrow indicates Dilp2 and DH44-GAL4 positive neuron. The other white regions in the image result from an overlap in z-projections between the two channels, rather than from antibody colocalization. E) Example membrane potential of a DH44PIN recorded in a fed (black) and a starved fly (orange). F) Baseline DH44PIN spike rate in fed (black) and starved flies (orange). Each circle represents an individual neuron (see table S6 for number of flies). Median and IQR are indicated. p-values are reported from Wilcoxon rank-sum test. G) Spike rate and delta spike rate of DH44PINs before and during glucose perfusion in starved flies. Plot details same as A). H) Schematic showing the location of cell bodies (left), and the regulation of IPCs, DH44PINs and DH44Ns outside the PI.

While IPCs and beta cells canonically sense glucose, the primary circulating sugar in the insect hemolymph is trehalose70,71, a disaccharide formed by two glucose molecules. Therefore, we tested whether IPCs can sense changes in the extracellular concentration of trehalose. To this end, we repeated our experiments with saline containing an equiosmolar concentration of trehalose (20 mM). As in previous experiments, we recorded the IPC baseline in glucose-free saline, which was also free of trehalose. Similar to glucose perfusion, we did not observe any significant changes in IPC activity during trehalose perfusion (Figure 3B), indicating that IPCs do not sense trehalose either.

These results clearly show that IPCs do not directly sense glucose or trehalose in their extracellular environment in vivo, but rather respond to the ingestion of glucose or other nutritive sugars (Figure 3C, see also Figure 1H). Hence, IPCs require input from the gut or other sensory pathways, such as taste or mechanosensation, to release insulin in response to glucose ingestion. Taken together, the increase in IPC activity following glucose ingestion but not glucose perfusion resembles the ‘incretin effect’ observed in mammals, where insulin secretion is significantly higher when glucose is ingested orally compared to isoglycemic intravenous delivery of glucose45,46. This effect is driven by gut-derived neuropeptides collectively referred to as ‘incretin hormones’ in mammals47,48. Recently, incretin-like hormones were also described in Drosophila72,73.

DH44PINs sense changes in extracellular glucose levels in vivo

We were surprised to find that IPCs were not responsive to glucose perfusion. To confirm that this was not due to shortcomings of our experimental protocol, we recorded from DH44PINs. Previous work showed that DH44PINs are activated by D-glucose via calcium imaging experiments74. DH44PINs are anatomically similar to IPCs, and their cell bodies are located directly adjacent to those of IPCs in the PI, making them an ideal positive control for our experiments (Figure 3D). A small subset of DH44PINs also expresses Dilp275, and our immunostainings confirmed colocalization of Dilp2 and DH44 in a single neuron (Figure 3D, white arrow).

When recording from DH44PINs in vivo, we first observed that their spike amplitude fell within the same range as that of IPCs (Figure 3E, ∼40-60 mV). However, DH44PIN baseline spike rates were higher than those of IPCs, ranging from 0.2 Hz to 3 Hz (Figure 3F). Moreover, DH44PINs exhibited a firing pattern that was more burst-like (Figure 3E), with shorter inter-spikeintervals (ISIs) than IPCs (Figure S4B). When comparing the ISI distribution between DH44PINs and IPCs, the fraction of ISIs below 500 ms was 60% in DH44PINs and 19% in IPCs. The cumulative ISI probability distributions differed significantly (Figure S4B, p << 0.001). Hence, the two neuron types displayed distinct activity patterns.

Next, we tested the effect of starvation on DH44PIN activity. The baseline spike rate was slightly lower in starved compared to fed flies, but, unlike in IPCs, this difference was not significant (Figure 3F). However, the cumulative probability distribution of DH44PIN ISIs differed significantly between starved and fed flies: ISIs were shifted towards longer durations after starvation (p << 0.001, Figure S4B). Hence, while the overall spike frequency was unaffected, and the baseline membrane potential remained unchanged (Figure S4C), DH44PIN spike patterns were more irregular and burst-like after feeding.

Finally, we tested whether DH44PINs are responsive to glucose perfusion. Following the exact same protocol we used for IPC recordings, we perfused the brain with saline containing 40 mM glucose, while patching from DH44PINs. The spike rate of DH44PINs significantly increased during perfusion with 40 mM glucose in starved flies, by about 0.5 Hz (Figure 3G). The ISI distribution was also significantly shifted towards shorter intervals during glucose perfusion across all flies and within each recording (Figure S4D and E). Fed flies, on the other hand, did not exhibit a significant change in DH44PIN activity during glucose perfusion (Figure S4F). This indicated that DH44PINs are indeed sensitive to glucose perfusion, and that the glucose sensitivity depends on the nutritional state, as previously reported74. Together, these experiments reveal two important points: First, DH44PINs are glucose sensing, whereas IPCs are not (Figure 3A). Second, the DH44PIN recordings confirm that that our negative result for IPC glucose sensitivity was not an experimental artifact.

DH44Ns outside the PI inhibit IPCs

The proximity of DH44PIN and IPC somata in the same cluster of the PI, their overlapping neurites in the brain (Figure 3D), and the fact that both populations are implicated in metabolic homeostasis prompted us to determine whether these neurons directly interact with each other. To investigate potential functional connectivity between DH44Ns and IPCs, we combined the UAS-GAL4 and LexA-LexAop systems to simultaneously express CsChrimson in all DH44Ns and GFP in IPCs (Figure 4A, see also Figure S5A)27. This approach enabled us to optogenetically activate DH44Ns while recording from IPCs via patch-clamp. Optogenetic activation of DH44Ns drove an immediate, strong hyperpolarization of IPCs, which lasted for > 1 s after stimulus cessation and reduced IPC spiking repeatedly across trials (Figure 4B). This effect was robust across the population of IPCs and flies (Figure 4C). On average, the IPC membrane potential dropped by –5 mV below the baseline during DH44N activation and stayed hyperpolarized until about 1 s after activation (Figure 4C). This fast and strong response suggests that DH44Ns could be directly presynaptic to IPCs. To quantify this effect, we calculated the median Vm for each IPC recording in a 500 ms window before the stimulus onset (pre, gray box in Figure 4C) and 200 ms after cessation of the stimulus (post, purple box in Figure 4C). All IPCs recorded were strongly hyperpolarized by DH44N activation (Figure 4D, p = 0.002). Hence, DH44Ns strongly and significantly inhibited IPCs. Notably, we also recorded from one IPC, which was excited during DH44N activation and inhibited immediately after cessation of activation (Figure S5C), indicating heterogeneity in the IPC population (see also Held et al.)30. Overall, our results indicate that acute activation of DH44Ns inhibits IPC activity, hence regulating insulin release. This strong inhibition is surprising, since our previous results implicated both, the DH44PINs and IPCs, in glucose sensing and metabolic control. Hence, our assumption had been that these modulatory neurons would display reciprocal excitation rather than inhibition. However, the driver line we used for DH44N activation was relatively broad, and any of the neurons labelled could be responsible for the IPC inhibition (Figure 4A). For example, the line labels a subset of DH44Ns ascending from the VNC. A pair of these DH44Ns co-express the neuropeptide Leucokinin (LK)76, and LK expressing neurons inhibit most IPCs30.

DH44Ns outside the PI inhibit IPCs and drive increases in locomotor activity

A) Immunolabelling showing DH44 expression in the brain and the VNC in the broad DH44-GAL4 driver line. GFP was enhanced with anti-GFP (green), brain and VNC neuropils were stained with anti-nc82 (magenta). B) Example recording of an IPC during optogenetic activation of the DH44Ns (red shading). Upper panel shows individual trials, lower panel shows ten trials overlaid and the median of all trials (purple trace). C) Upper panel shows the spike density of individual IPCs across 10 DH44N activations. Lower panel shows the baseline-subtracted, median filtered Vm traces for each IPC. Thick lines represent the grand mean. D) Effect of DH44N activation on IPCs. Delta Vm is plotted by calculating the median baseline subtracted Vm from C) 500 ms before (Pre) and 200 ms after DH44N activation (Post). Each circle represents one IPC recording. p-values from Wilcoxon signed-rank test. E) Immunolabelling showing GFP expression in the brain and the VNC in the sparse DH44PI-GAL4 driver line. F) Example recording of an IPC during optogenetic activation of DH44 neurons using the sparse DH44PIN driver line. Plot details as in B). G) Spike density and baseline subtracted median Vm of individual IPCs during activation of the DH44PIN driver line. Plot details as in C). H) Pre and post delta Vm of IPCs before and after optogenetic activation of the DH44PIN line. Plot details as in in D). See table S5 for number of flies. I) Average FV of 20 flies during optogenetic activation of the DH44Ns using the broad driver line (see A). Empty split-GAL4 was used as a control for all experiments (black). Red shading indicates 300s optogenetic activation windows. J) Average FV across all DH44N activation trials based on two independent replications of the experiment in I. Note that the peak in average FV lies within the first frame of the stimulation window. K) Average FV of all flies during each stimulus trial (1-5) and post-stimulus trial (300 s window immediately after activation ceased, P1-P5). Circles and bars show median and IQR, respectively. Asterisks represent a significant difference according to a Wilcoxon rank-sum test. L-N) Behavioral effects of optogenetic DH44PIN activation (see E). Plot details as in I, J and K, respectively. See Table S3 and S4 for p-values.

To investigate this, we identified a driver line which only labels the DH44Ns in the PI77 (DH44PI–GAL4) and repeated the functional connectivity experiments using this more specific driver line (Figure 4E, see also Figure S5B). Activating exclusively DH44PINs had no effect on the IPC membrane potential or spike rate (Figure 4F and, 4G: grand mean remained 0 mV). This was consistent across the population of IPCs and flies (Figure 4H, p = 0.641). These results confirm that the strong inhibition of IPCs observed when activating the broad DH44N line is not driven by DH44PINs. Therefore, DH44PINs and IPCs both sense change in the nutritional state – IPCs after feeding and DH44PINs via hemolymph sugar concentrations – but act independently and in parallel to signal these changes (Figure 3H).

Activation of DH44Ns modulates locomotor activity

Given their sensitivity for glucose and connectivity to IPCs, DH44Ns were strong candidates to contribute to the effects of nutritional state changes on locomotor activity (Figure 2). DH44Ns were previously shown to promote feeding in starved animals74. In particular, we hypothesized that the broad DH44N line could have opposite effects to IPCs since a subset of them inhibited IPCs. We therefore tested whether DH44Ns affect locomotor activity by following the same approach as for IPC and OAN activation in freely walking flies (Figure 2C-K). Activation of the broad DH44N line resulted in a dramatic increase in FV from a median of 1.8 mm/s to 6 mm/s immediately after stimulus onset (Figure 4I, 4J). This increase was followed by an abrupt halt that lasted until DH44N activation ceased (Figure 4I-K). During this time window of reduced locomotor activity, we observed a strong increase in proboscis extension (PE) in 6/20 flies for which we quantified this (Figure S5D). The locomotor arrest was released soon after cessation of DH44N activation, upon which locomotor activity increased until the onset of the next stimulation period (Figure 4I and 4K). The FV in the post-activation window increased over subsequent activation cycles and was significantly higher than in controls throughout the experiment (Figure 4I, 4K). The FV of DH44N activated flies increased from a median of 0.3 mm/s to 3 mm/s over five activation cycles, suggesting that long-term activation of DH44Ns leads to an increase in locomotion comparable to starvation-induced hyperactivity.

In addition, we tested the driver line labelling only DH44PINs. Although we did observe an overall increase in the FV of the experimental group, we neither observed an activity peak upon activation onset, nor locomotor arrest during activation (Figure 4L-N). Furthermore, contrary to activation of the broad DH44 driver line, specific activation of the DH44PINs did not drive PE (Figure S5E). Hence, the strong activation phenotypes were driven by DH44Ns outside the PI. We conclude that the activation of DH44Ns, especially the broad DH44N line, had a pronounced and sustained effect on foraging behavior. The same neurons strongly and significantly inhibited IPCs, suggesting that feeding-promoting and satiety-inducing pathways form inhibitory connections in the CNS as a part of the neuronal mechanisms to ensure reciprocal inhibition of satiety and hunger states.

Discussion

IPC activity is modulated on short and long timescales by the nutritional state, the behavioral state, and aging processes

To advance our understanding of the neuronal underpinnings of metabolic homeostasis, we analyzed the nutritional state-dependent modulation of IPCs in Drosophila. Our in vivo patch clamp recordings revealed that IPC activity is diminished during starvation and rebounds after glucose feeding. This is consistent with previous studies demonstrating an accumulation of DILPs26,78–82 and a reduction of dilp transcript levels79,81,82 in IPCs of adult and larval fly brains under starvation and underlines that electrophysiological IPC activity and DILP release are correlated. However, two aspects were surprising about our results. First, the rebound after glucose feeding was relatively slow, and IPC activity built up to baseline levels after over six hours of refeeding. These slow temporal dynamics are in stark contrast to the behavioral state-dependent modulation of IPC activity, which occurs on a timescale of milliseconds27. This suggests that IPCs integrate changes in internal states over different timescales and potentially affect metabolism and neuronal networks on both short and long timescales. The second surprising result was that the reduced IPC activity after starvation failed to rebound upon refeeding with high protein and standard diets (Figure 1H). Our standard diet contains glucose at very low concentrations (upper bound of 30 mM, see Methods for details), which partially explains the lack of an IPC activity rebound. However, the observation that IPC activity in flies refed with SD remained comparable to that of starved flies suggests that starvation had a long-lasting effect on IPC activity, which could specifically be restored by feeding nutritive sugars (glucose and fructose). One potential explanation could be that older flies exhibited lower IPC activity levels, which could lower the baseline activity after refeeding and thus counteract the rebound in IPC activity. However, we always recorded from flies aged between 5-6 days in refeeding experiments, and a significant decrease in IPC activity was only observed in flies older than 8 days (see Figure S2B). This suggests that aging played only a minor part here. Instead, reduced IPC activity after refeeding with SD may indicate increased insulin sensitivity as a result of the starvation-refeeding paradigm. Increased insulin sensitivity has been reported in multiple mammalian studies which used a starvation-refeeding paradigm to study effects of dietary restriction on insulin release83,84, and our results indicate that this could be similar in flies. DILP6 mRNA in the fat body is upregulated during starvation, which has been shown to decrease systemic insulin signaling by the suppression of insulin release from IPCs80, potentially contributing to mechanisms regulating insulin sensitivity in flies. Furthermore, we observed that IPC activity did not recover on the HP diet, which also resulted in poor survival rates. This may indicate that insulin signaling is impaired in flies if they only feed on protein. Similar observations were made in mammals, where prolonged high-protein diets lead to insulin resistance and type II diabetes82,85.

One of our experiments demonstrated that IPC activity was heavily diminished in flies older than 10 days (Figure S2B). A possible explanation could be that flies feed less as they age. However, this only holds true for flies older than 14 days86. Therefore, reduced IPC activity in 10-11 day old flies is unlikely to result from reduced food intake and likely involves inhibition of insulin signaling. For instance, dilp2 mutants show an increased lifespan87 which could partially result from elevated glycogen phosphorylase activity88, an enzyme involved in regulating glycogen storage. This suggests that suppressed IPC activity in aged flies could mediate glucose homeostasis by increasing glycogen catabolism. Reduced insulin signaling has previously been linked to increased life-span in different species, including yeast, worms, flies, and rodents1. The suppression of IPC activity could be a mechanism to increase survival under physiologically unfavourable conditions. In summary, the reduced IPC activity we observed in starved and aged flies may be crucial for metabolic adjustments required for survival. Notably, similar mechanisms have been observed in mammals, where reduced insulin signaling and increased insulin sensitivity during fasting or dietary restriction contribute to enhanced metabolic efficiency and longevity. These parallels suggest that the regulation of insulin signaling in response to nutrient availability is an evolutionarily conserved strategy, underscoring its fundamental role in survival and metabolic homeostasis across species.

The nutritional state-dependent modulation of IPCs supports the presence of an incretin-like effect in Drosophila

Mammalian pancreatic beta cells sense glucose levels in the blood and release insulin accordingly89,90. However, this process is much more complex than initially assumed. One key observation is that the ingestion of glucose drives much higher release of insulin compared to the isoglycemic intravenous perfusion of glucose. This difference in insulin release is partially driven by incretin hormones from the gut, for example glucagon-like peptide-1 (GLP-1) and glucose-dependent insulinotropic peptide (GIP)91–93, which are released after the ingestion of food. Hence, insulin release is modulated by enteroendocrine signals. Drosophila IPCs are functionally analogous to mammalian beta cells and presumably sense glucose in the circulating hemolymph to release insulin in response to increases in glucose concentration. Previous ex vivo studies suggested that IPCs, like pancreatic beta cells, sense glucose cell-autonomously23,44. Consistent with this, we observed an increase in IPC activity after the ingestion of glucose (Figure 2B). However, IPC activity did not increase during the perfusion of glucose directly over the brain. Importantly, the fly preparations were kept alive for several hours allowing the glucose-rich saline to enter circulation and reach all body parts. Several factors may explain the difference between ex vivo and in vivo preparations. First, in ex vivo studies, certain regulatory feedback mechanisms present in vivo could be absent. For example, the strong inhibitory input IPCs receive from DH44Ns we found would likely be absent in brain explants without a VNC. A lack of inhibitory feedback might allow for more direct glucose sensing by IPCs ex vivo, whereas in vivo, the IPC response could be suppressed by more complex systemic feedback. Second, we attempted to use the intracellular saline formulation employed in a previous ex vivo study44. However, we observed that IPCs depolarized quickly using this saline, leading to unstable recordings that did not meet our quality standards for in vivo experiments. Another possible explanation for the lack of an effect of glucose might have been that the dominant circulating sugar in flies is trehalose70,71 which is derived from glucose. When we extended our experiments, we found that trehalose perfusion did not affect IPC activity either, strengthening the idea that IPCs do not directly sense changes in hemolymph sugar levels. Therefore, our findings suggest that, similar to mammals, IPC activity and hence, insulin release, is not simply modulated by hemolymph sugar concentration in Drosophila. Several pathways could contribute to this. First, although an incretin effect has not been reported in flies before, recent studies provided evidence of incretin-like hormones such as the midgut-derived neuropeptide F in flies72,94, suggesting that a similar signaling pathway could underlie the incretin effect across species. Second, a subset of IPCs was shown to express the mechanosensitive channel piezo and innervate the crop95, suggesting that an increase in crop volume during feeding could contribute to the increase in IPC activity after feeding. However, the fact that we observed strong responses of IPCs to glucose intake but not the ingestion of regular food during refeeding suggests this is not the core mechanism underlying the incretin-like effect we observed. Third, taste pathways converge onto IPCs96 and they could increase IPC activity after the ingestion of glucose via serotonergic inputs from taste responsive cells in the gnathal ganglion. Here, however, we would expect faster responses on the scale of seconds, which are unlikely to shift IPC activity over several hours. Hence, gut-secreted neuropeptides are the strongest candidates to mediate the incretin effect in Drosophila. These results signify a causal link between glucose homeostasis in the gut and the brain and pave the way for further studies investigating the evolutionary conservation of the incretin effect across mammals and insects.

We observed that IPC activity increased over a timescale of hours, which is longer compared to the fast insulin response in mammals, where insulin typically peaks within an hour of feeding97. In flies, insulin levels rise within minutes of refeeding, followed by a drop after 30 min20. Our experimental techniques limit our ability to capture these fast initial dynamics, since the preparation for intracellular recordings requires tens of minutes, so that we typically recorded IPC activity at least 20 min after the last food ingestion. Notably, studies in fasted mammals have shown that insulin peaks within minutes of refeeding, followed by a rapid decline, with levels stabilizing as feeding continues98,99. We speculate a similar dynamic could be present in flies, but with our approach, we capture the steady-state reached tens of minutes after food ingestion rather than a potential initial peak.

DH44PINs and IPCs have distinct, yet interrelated roles in nutrient sensing

Our in vivo recordings revealed that IPCs do not sense glucose cell-autonomously, and neither do neurons in the CNS that are directly connected to IPCs – otherwise we would have expected an increase in IPC activity during glucose perfusion. In contrast, DH44PINs were sensitive to changes in the extracellular glucose concentration. This could either be driven by cell-autonomous glucose sensitivity or presynaptic glucose sensing neurons. Moreover, DH44PINs did not show an increase in activity in fed flies, indicating a lack of sensitivity to glucose ingestion. IPCs responded to glucose ingestion on a timescale of hours, whereas DH44PINs immediately responded to increases in extracellular glucose concentrations on a timescale of seconds to minutes. These results suggest that DH44PINs and IPCs play distinct roles in glucose sensing, which might complement each other. Our initial functional connectivity experiments suggested that the glucose-sensing DH44PINs might inhibit IPCs directly. However, refined experiments using a sparse driver line revealed that IPCs are inhibited by DH44Ns outside the PI. If these DH44Ns, like DH44PINs, are glucose sensing they would suppress IPC activity in the presence of high glucose concentrations, which would be counteractive. Therefore, even though this remains to be determined, it is likely that these non-PI DH44Ns are activated by signals other than glucose concentrations. Notably, the DH44PINs express the DH44 peptide, as confirmed by anti-DH44 stainings100. This also applies to a large fraction of neurons labelled in the broad DH44 driver line100. However, a subset of neurons labelled in the broad line did not exhibit DH44 immunoreactivity100, and might therefore not actually express the DH44 peptide. Hence, the inhibition of IPCs could be driven by neurons in the DH44 driver line that do not express DH44. A strong candidate for the inhibition are LK and DH44-positive neurons, which are labelled by the broad line76. In a parallel study, we showed that LK-expressing neurons strongly inhibit IPCs30, similar to the broad DH44 line used here. Furthermore, evidence from single-nucleus transcriptomic analysis shows that IPCs express DH44-R1 and DH44-R2 receptors30. Therefore, it is possible that DH44Ns communicate with IPCs through a direct peptidergic connection. Notably, the inhibitory effect of non-PI DH44Ns on IPCs was very strong and fast, suggesting that a connection via classical synapses is more likely. Regardless, our results show that the glucose sensing DH44PINs and IPCs act independently of each other.

In line with previous studies74,101, we found that DH44PINs were only activated by glucose in starved flies. Interestingly, this was not because the DH44PIN spike rate was lower after starvation. Instead, DH44PINs exhibited similar spike rates in fed and starved flies, but the activity increase in response to glucose perfusion only occurred under starvation. DH44PINs also displayed a level of modulation beyond simple increases and decreases in spike frequency. In particular, the ISI distributions of DH44PINs were shifted to significantly longer intervals during starvation. This may contribute to the increased glucose sensitivity in starved flies. Moreover, shorter ISIs in fed flies and the more bursty activity pattern could lead to an increase in DH44 peptide release despite similar overall spike frequencies. Apart from this, DH44PINs themselves could also become more sensitive to glucose in starved animals. This would suggest that DH44PINs sense changes in glucose concentration, rather than the absolute concentration of glucose, in the brain or hemolymph.

Nutritional state-dependent modulation of behavior

Changes in the activity of modulatory neurons affect information processing in neuronal circuits and ultimately behavior22,55,102–104. This includes starvation-induced hyperactivity and other aspects of foraging in Drosophila. In this regard, we first tested whether IPCs modulate locomotor activity in flies. This was prompted by the observation that starvation strongly modulated both IPC activity and locomotion. However, IPCs only had a minor effect on locomotion. This was despite stimulating IPCs using long, highintensity optogenetic activation. The strongest effect we observed was a reduction in starvation-induced hyperactivity upon IPC activation, which we interpret as a satiety-like effect due to increased IPC activity. This effect was stronger in the beginning of the experiment, even before the first activation of IPCs, and was robust across independent repetitions of the same experiment. This was likely in part due to an increased IPC depolarization via CsChrimson activation because of exposure to white light during the preparation, recovery period, and behavioral experiment. We carried out all behavioral experiments in low intensity white light. Background light levels strongly affect locomotor activity in flies105, and overall we observed more reliable starvation-induced hyperactivity under light conditions. We suspect the optogenetic effect of background light was small, though, because in genotypes that had very strong and acute effects on locomotion under red light the baseline locomotor activity under background light conditions was in the range of controls (Figure 2I-K and Figure 4I-K). Another reason why we performed experiments under light conditions and not in the dark is that switching from dark to light leads to a startle response in optogenetic activation experiments when the red LED is switched on, which masks behavioural responses induced by neuronal activation. We minimized this effect by using low intensity white background light. With subsequent activations, the difference between IPC-activated and control flies was reduced, probably because flies were increasingly food and water-deprived after an hour of activation experiments. In particular the latter could potentially override the reduction in starvation-induced hyperactivity induced by optogenetic IPC activation. Overall, our experiments revealed that IPC activation and resulting insulin release do not lead to a sustained satiety response observable in behavior. This aligns with physiological expectations, as elevated insulin levels exacerbate the glucose deficit in starved flies, which could ultimately lead to increases in hyperactivity via this secondary effect.

In contrast to insulin, we confirmed the causal role of one of the known neuromodulators involved in driving starvationinduced hyperactivity – octopamine. As expected, OAN activation increased locomotor activity substantially, which we interpret as a proxy for foraging based on previous publications38,106. The activity levels reached by long-term OAN activation were comparable to the effects of long-term starvation. These results underscore that octopamine plays a major role in driving starvationinduced hyperactivity. Our results also suggest a key role of DH44Ns in modulating aspects of foraging behavior. There is evidence that DH44 release in the brain modulates rest-activity rhythms107 via downstream targets in the gnathal ganglion108, which houses circuits involved in the modulation of feeding and locomotion34,108–113. Our optogenetic approach provides substantial evidence that DH44Ns drive increased locomotor activity in adult flies, plausibly to enable exploration of the environment for locating food sources. First, the increase in locomotion phenocopies starvation-induced hyperactivity. Second, our experiments additionally showed that DH44Ns outside the PI drive proboscis extensions during stopping, which are a hallmark of feeding behavior. In line with this, a previous study has shown that activation of DH44 receptor 1 neurons, the target neurons of the DH44 neuropeptide, induces proboscis-extension74. Both observations suggest that DH44Ns are a component of feeding pathways and mediate nutrient uptake in order to maintain nutritional homeostasis.

Our functional connectivity experiments revealed that these DH44Ns inhibit IPC activity. At first glance, promoting feeding and inhibiting insulin release at the same time seems counter-regulatory. However, there are two possible explanations: First, these two effects could be driven by different sub-populations of DH44Ns labelled by the line we used (Figure 4A and E). Second, the IPC inhibition could be part of a combination of physiological effects signaling a hunger state, and thus contribute to the increased foraging activity driven by DH44Ns. Notably, blocking insulin signaling has been reported to mimic starvation in fed flies55. Contrary to DH44Ns outside the PI, activation of the glucose-sensing DH44PINs did not drive proboscis extension, suggesting that DH44PINs are not involved in feeding initiation, but primarily function in post feeding regulation of metabolic homeostasis. This is in line with our electrophysiological results showing that DH44PIN spike frequency is not affected by starvation, and only starts to increase once glucose levels change in the hemolymph. Furthermore, DH44PIN activation slightly increased the FV but did not lead to stopping, suggesting that DH44PINs contribute to starvation-induced hyperactivity, whereas other DH44Ns drive stopping and proboscis extension.

In conclusion, our study sheds light onto the intricate connections between neuroendocrine signaling, nutrient sensing, and behavior in Drosophila. Using electrophysiological approaches, we unraveled the complex activity dynamics of IPCs and DH44PINs and identified differences in their activity patterns during various nutritional states. The discovery of an ‘incretin-like’ effect in flies suggests that important aspects of gut-brain signaling are conserved across vertebrate and invertebrate species. Hence, our findings not only contribute to the increasing body of research on insulin signaling, but also pave the way for future research analyzing the intricacies of Drosophila neuroendocrine networks governing nutrient sensing and behavior.

Materials and methods

Fly husbandry

The following fly strains were used: DILP2-GAL4; CyO (BDSC #37516), 10XUAS-IVS-myr::GFP (BDSC #32197), TDC2-GAL4 (#BDSC 9313), DH44-GAL4 (BDSC #39347), DH44PI-GAL4 (BDSC #51987), R96A08-LexA-p65-vk37::LexOp-dilp2::GFP; 20x-UAS-CsChrimson-attp2/TM6b (as described in Liessem et al.)27, 10XUAS-GFP-P10 (BDSC #32201), 20x-UAS-CsChrimson (BDSC #55134), Empty split-GAL4 (BDSC #86738). DILP2-GAL4 and DH44-GAL4 driver lines were crossed with 10XUAS-IVS-myr::GFP to drive the expression of GFP in IPCs and DH44Ns, respectively, for patch clamp recordings. For electrophysiology combined with optogenetics, DH44N driver lines were crossed with R96A08-LexA-p65-vk37::LexOp-dilp2::GFP; 20x-UAS-CsChrimson-attp2/TM6b, hence expressing CsChrimson in DH44Ns and GFP in IPCs.

Flies were raised at 25°C and 60% humidity under a 12h/12h light/dark cycle, on standard fly food containing per liter: 147.5 g cornmeal, 10 g soy flour, 18.5 g Cenovis (beer yeast), 6.25 g agar-agar, 45 g malt syrup, 45 g sugar beet molasses, and 2.5 g Nipagin. For starvation, flies were kept in empty vials with a filter paper soaked in water. For refeeding experiments, starved flies were refed on one of the following diets: 400 mM D-glucose in 2% agar (HG), 400 mM D-glucose in standard diet (HG+SD), 400 mM D-fructose in 2% agar (HF), 400 mM D-arabinose in standard diet (HA+SD), 10% yeast extract in 2% agar (HP) or standard fly food (standard diet). For optogenetic experiments, 300 µM of all-trans-retinal (R2500, Sigma-Aldrich, Steinheim, DE) was added to 10 ml of fly food. These vials were kept in darkness until the flies were used for experiments. Our standard lab diet does not contain glucose as such, but we estimated the concentration of glucose based on indirect sources of glucose (malt syrup and sugar beet molasses). Based on the percentage of glucose present in malt syrup (7%) and sugar beet molasses (4.55 g per 100 g), we estimate that our lab diet contains about 28 mM of glucose.

Electrophysiology

All experiments were performed in mated females between 3-6 days post eclosion, except for comparison of IPC activity between different age groups and mating states. Flies were cold anesthetized on ice and then immobilized in a custom-made shim plate fly holder using UV glue (Proformic C1001, VIKO UG, Munich, DE). The proboscis was glued to the thorax to restrict brain movement, and the front legs were excised. The cuticle was then removed in a window above the pars intercerebralis so that the posteriordorsal part of the fly brain was exposed. Trachea and the ocellar ganglion were removed to render the IPCs and DH44PINs accessible. The fly was then transferred with the fly holder into a customized, upright fluorescence microscope setup (Olympus BX51WIF, Evident Corporation, Tokyo, JPN). Live images of the fly brain were acquired with a high-resolution camera (SciCam Pro, Scientifica, Uckfield, UK) using an image acquisition software (OCULAR™, Digital Optics Limited, Auckland, NZ).

During the preparation, as well as the experiment, the brain was continuously perfused with carbonated (95% O2 and 5% CO2) extracellular saline containing, 103 mM NaCl, 3 mM KCl, 5 mM N-[Tris(hydroxymethyl)methyl]-2-aminoethanesulfonic acid, 20 mM sucrose, 26 mM NaHCO3, 1 mM NaH2PO4, 1.5 mM CaCl2.2H2O, 4 mM MgCl2.6H2O, and osmolarity adjusted to 273-275 mOsm (modified from Gouwens & Wilson, 2009)114. 0.025% Collagenase (w/v in extracellular saline, Sigma-Aldrich #C5138) was gently applied using a thin-walled glass pipette to dissociate the neural sheath above cell bodies before recordings. IPC/DH44PIN cell bodies were identified via GFP expression and whole-cell patch clamp recordings were performed using thickwalled patch pipettes (4 – 8 MΩ resistance) containing intracellular saline (40 mM potassium aspartate, 10 mM HEPES, 1 mM EGTA, 4 mM MgATP, 0.5 mM Na3 GTP, 1 mM KCl and 20 μM, 265 mOsm, pH 7.3). Continuous time series recordings of the membrane potential were captured in current clamp mode with an AxoPatch MultiClamp 200B (Molecular Devices, Sunnydale, CA, USA) and corrected for a 13 mV liquid junction potential114. All data were recorded with a Digidata 1440A analog-digital converter (Molecular Devices), controlled by the pCLAMP 10 software using a 10 kHz low-pass filter and a 20 kHz sampling rate. Recordings were accepted for analysis if the resting membrane potential was < –48 mV and the spike amplitude > 20 mV.

Baseline activity was always recorded in glucose-free saline for 10 min, and the last five min were used for analysis. For glucose perfusion experiments, IPC activity was recorded in glucose-free extracellular saline for 10 min to establish the baseline. Next, the brain was perfused with extracellular saline replacing 40mM sucrose with 40 mM glucose. Similarly, trehalose-rich saline was prepared by substituting 40 mM sucrose with 20 mM trehalose to maintain consistent osmolarity. The switch between the salines took about three min, at which point the glucose/trehalose-rich saline reached the brain. The membrane potential was then recorded for 15 min in three consecutive five min recordings. Out of these, the last two recordings (Glu1/Tre1 and Glu2/Tre2) were used for analysis to make sure that glucose had time to reach the IPCs and deeper structures inside the brain.

For optogenetic activation experiments during electrophysiological recordings, CsChrimson was activated using a 625 nm LED with an intensity adjusted to ∼4.4 mW/cm2 at the position of the fly. The activation protocol was triggered by the pCLAMP software. Each recording consisted of 10 activations with the LED ON for 100 ms, interleaved with a 10 s long inter-stimulusinterval. TTL triggers were used to drive the LED and were recorded simultaneously with intracellular traces in the abf format (Axon Binary File).

Quantification and statistical analysis for electrophysiology

All data analysis was performed in MATLAB R2021a (The Mathworks, Natick, MA, USA). The statistical tests we used, and respective p-values are mentioned in the figure legends, captions, and supplementary tables. Data points are presented with median and interquartile ranges unless otherwise noted.

Intracellular recordings were temporally smoothed using the smooth function in MATLAB and the ‘loess’ method with span 7 at a 20.000 Hz sampling rate. Spike thresholds were set manually in either the recording or the derivative of the recording to detect spikes. Accurate spike detection was monitored manually. For the analysis of changes in membrane potential (Vm), the median of intracellular traces was used. Spike density of individual neurons across the 10 different trials of the short stimulation protocol was computed by using 250 ms time bins.

Immunohistochemistry and image acquisition

For immunolabelling, respective driver lines were crossed with a UAS-GFP-P10 or UAS-myr-GFP reporter to express GFP in the neurons of interest. Flies were anesthetized on ice and then transferred to 1.5 ml 4% paraformaldehyde in 0.1 M phosphate buffered saline containing 0.5% Triton-X100 (PBT). The flies were fixed in this solution for 3-6 h on a shaker followed by three 15 min wash steps in PBT. Individual fly brains were then dissected in a SYLGARD® dish after immersing the fly briefly in 70% ethanol. Dissected brains were then washed three times in PBT for 15 min followed by blocking in 10% normal goat serum in 0.1 M PBT (blocking buffer), for either 2 h at room temperature or overnight at 4°C. Next, the samples were incubated for 2 days at 4°C in primary antibody solutions in blocking buffer. We used rabbit anti-DILP2 diluted to 1:2000 (RRID: AB_2569969, kindly provided by J. A. Veenstra, Bordeaux, FR) to label IPCs, mouse anti-nc82 (Bruchpilot C-terminal aa 1227-1740)115 diluted as 1:500 to label the neuropils as background, and chicken anti-GFP (ab13970, abcam, Berlin, DE) diluted to 1:1000 to enhance the endogenous GFP signal. Afterwards, brains were washed in PBT for three consecutive 15 min washes, followed by overnight incubation in secondary antibody solutions at 4°C in the dark. Here, we used AlexaFluor-488 (goat anti-chicken IgY (H+L), Thermo Fisher Scientific, Waltham, MA, USA) diluted to 1:200, AlexaFluor 635 (goat anti-mouse IgG (H+L), Thermo Fisher Scientific) diluted to 1:400, and AlexaFluor 555 (goat anti-rabbit IgG (H+L), Thermo Fisher Scientific) diluted to 1:200. Finally, samples were washed three times in PBT before mounting them in Vectashield Antifade Mounting Medium (H-1000, Vector Laboratories, Newark, CA, USA).

Image acquisition was performed with a confocal laser-scanning microscope (Leica TCS SP8 WLL, Leica Microsystems, Wetzlar, DE) through the Leica Application Suite (LAS X, Leica Microsystems) using HC PL APO 10×/0.4, HC PL APO 20×/0.75 IMM, and HC PL APO 63×/1.2 CORR objectives. Fluorescence was detected using suitable lasers with a resolution of 1024 x 1024 pixels, in serial stacks. Final image processing to adjust contrast and brightness was done in Fiji116.

Survival assay

For conducting the survival assay, two replicates of 20 flies per condition were used. Initially, 3-4 days old adult female flies were wet-starved for 24 h, and then refed on either HG, HP, or SD. They were kept on these diets for the rest of their survival span and were transferred onto fresh food every second day. The number of survivors was counted once per day and dead flies were removed. For analysis of survival percentage, we combined the counts from two replicates.

Free-walking assay

Virgin female flies carrying 20x-UAS-Cs-Chrimson-m-venus58 were mated with males from the neuromodulator GAL4 lines studied. Post eclosion, the offsprings were reared on standard food coated with 300 µM all-trans retinal for a period of three to six days prior to the experiment. In the experiments that involved starved flies, flies were transferred to a plastic vial containing a small piece of wet tissue paper for 24 h before the start of the experiment. The walking activity of up to 20 flies was recorded using the ‘Universal Fly Observatory’ (UFO) behavioral setup which is based on an earlier study56. The UFO features a walking arena consisting of an inverted petri dish of 100 mm diameter as its base. To prevent escape, this base was covered with a watch glass of 120 mm diameter. The inside of the watch glass was coated with SigmaCote siliconizing reagent (cat. no. SL2; RRID:SCR_008988, Sigma-Aldrich, St. Louis, MO, USA) to prevent flies from walking upside down on the watch glass. The arena was illuminated by a ring of 60 infrared (IR) LEDs (wavelength: 870 nm) arranged concentrically around the arena. A diffuser was placed along the inside of the IR LED ring to ensure uniform illumination across the entire arena. The arena was filmed from below via a surface mirror positioned at a 45° angle below the arena and a camera (Basler acA1300-200um, Basler AG, Ahrensburg, DE) with a spatial resolution of 0.08 mm/pixel and a temporal resolution of 20 Hz. The camera’s lens was equipped with an IR-pass filter (lower cutoff frequency: 760nm) that only permits the IR light to pass through to ensure a strong contrast between the background (black) and flies (bright). A second ring of 60 red LEDs (wavelength: 625 nm) positioned above the IR LED ring provided the light necessary to transiently activate CsChrimson for optogenetic activation. The entire UFO setup was enclosed within a box, together with a second UFO. The box included a visible white light source placed above the arenas for controlled background light conditions.

Before each experiment, twenty flies were transferred to an empty vial and anesthetized on ice for two min. Subsequently, flies were transferred to the UFO arena. They were given a recovery period of 10 min to explore the arena before starting the experiment. One complete experiment consisted of five activation cycles and each activation cycle lasted 15 min and included 5 min of exposure to red light at an intensity of 3.19 mW/cm2, for a total duration of 90 min per experiment. All experiments were conducted under low brightness (0.112 mW/cm2) white light conditions. For each neuromodulator line, the experimental flies and control flies were recorded simultaneously in two separate UFOs situated in the same enclosure. Both the protocol and video acquisition were controlled with custom-written MATLAB code.

Movement of flies was tracked using the Caltech Fly Tracker software57. A detailed description of the tracking software is available at https://kristinbranson.github.io/FlyTracker/index.html. The position of individual flies was extracted from the videos on a frame-by-frame basis and features like translational and angular velocity were further computed in MATLAB. A 2 s median filter, which corresponds to less than 1% of the 300 s stimulation window, was applied prior to analysis. We excluded flies that died during the experiments from the tracking dataset prior to analysis. We conducted a minimum of 2 replicates for each experimental line and its corresponding controls and combined their data for further analysis. For plotting the average velocity across all flies, we implemented a data down-sampling by a factor of 20. Statistical analysis was conducted using MATLAB with the Wilcoxon rank-sum test.